---
title: "CA vs FL Homelessness EDA Graphs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages and data

```{r load}
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)

df <- read_excel("CA_FL_LASSO_MODEL_INPUT_v2.xlsx", sheet = "LASSO Model Data")
```

## Graph 1: Raw target distribution
```{r graph_2021_optional, eval=FALSE, include=FALSE}
# --- add the 2021 point back in ---
# 2021 is not in `df` because it is excluded as a modeling target.
# This pulls it from the separate descriptive dataset, for plotting only.
df_2021 <- read.csv("outputs/descriptive_2021/coc_homeless_rate_by_year_INCLUDING_2021.csv",
                    stringsAsFactors = FALSE) %>%
  filter(year == 2021) %>%
  transmute(state, coc_number,
            target_year                  = year,
            target_homeless_rate_per_10k = homeless_rate_per_10k)

df_plot <- bind_rows(
  df %>% select(state, coc_number, target_year, target_homeless_rate_per_10k),
  df_2021)

state_trends <- df_plot %>%
  group_by(state, target_year) %>%
  summarise(mean_rate = mean(target_homeless_rate_per_10k), .groups = "drop")

# solid line skips 2021 (the trend as the models see it);
# dashed detour shows where the disrupted 2021 point actually sits
trend_solid  <- state_trends %>% filter(target_year != 2021)
trend_dashed <- state_trends %>% filter(target_year %in% 2020:2022)

ggplot(df_plot, aes(x = target_year, y = target_homeless_rate_per_10k, color = state)) +
  annotate("rect", xmin = 2020.5, xmax = 2021.5, ymin = -Inf, ymax = Inf,
           fill = "grey60", alpha = 0.18) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1) +
  geom_line(data = trend_dashed, aes(x = target_year, y = mean_rate, color = state),
            linewidth = 0.9, linetype = "dashed") +
  geom_line(data = trend_solid, aes(x = target_year, y = mean_rate, color = state),
            linewidth = 1.2) +
  geom_point(data = trend_solid, aes(x = target_year, y = mean_rate, color = state),
             size = 2.5) +
  geom_point(data = state_trends %>% filter(target_year == 2021),
             aes(x = target_year, y = mean_rate, color = state),
             size = 3, shape = 21, fill = "white", stroke = 1.2, show.legend = FALSE) +
  annotate("text", x = 2021, y = max(df_plot$target_homeless_rate_per_10k) * 0.99,
           label = "2021\nCOVID-disrupted count", size = 3, colour = "grey25",
           lineheight = 0.95) +
  scale_color_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  scale_x_continuous(breaks = seq(2012, 2025, 2)) +
  labs(title = "CA vs FL: next-year homeless rate over time",
       subtitle = paste("Lines = UNWEIGHTED mean across CoCs; points = individual CoC-years.",
                        "\n2021 is hollow: 47 of 71 CoCs counted sheltered people only, so the dip is a counting artifact."),
       x = "target_year", y = "homeless rate per 10k") +
  theme_minimal()
getwd()

list.files(
  recursive = TRUE,
  pattern = "\\.csv$",
  full.names = TRUE
)
csv_path <- "replace/with/the/correct/file.csv"

file.exists(csv_path)
```

```{r graph1}
skew <- function(x) {
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  sum((x - m)^3) / n / s^3
}

skew_raw <- round(skew(df$target_homeless_rate_per_10k), 2)

ggplot(df, aes(x = target_homeless_rate_per_10k)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Target distribution: next-year homeless rate per 10k",
       subtitle = paste0("n=", nrow(df), ", skewness=", skew_raw),
       x = "target_homeless_rate_per_10k", y = "density") +
  theme_minimal()
```

## Graph 2: Log target distribution

```{r graph2}
skew_log <- round(skew(log(df$target_homeless_rate_per_10k)), 2)

ggplot(df, aes(x = log(target_homeless_rate_per_10k))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Log target distribution",
       subtitle = paste0("skewness=", skew_log),
       x = "log(target_homeless_rate_per_10k)", y = "density") +
  theme_minimal()
```

## Graph 3: CA vs FL target trends over time

```{r graph3}
state_trends <- df %>%
  group_by(state, target_year) %>%
  summarise(mean_rate = mean(target_homeless_rate_per_10k), .groups = "drop")

ggplot(df, aes(x = target_year, y = target_homeless_rate_per_10k, color = state)) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1) +
  geom_line(data = state_trends, aes(x = target_year, y = mean_rate, color = state),
            linewidth = 1.2) +
  geom_point(data = state_trends, aes(x = target_year, y = mean_rate, color = state),
             size = 2.5) +
  scale_color_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "CA vs FL: next-year homeless rate over time",
       x = "target_year", y = "homeless rate per 10k") +
theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(size = 19, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 14)
  )
```

## Graph 4: CoC-level target boxplots by state, over time

```{r graph4}
ggplot(df, aes(x = factor(target_year), y = target_homeless_rate_per_10k, fill = state)) +
  geom_boxplot(outlier.size = 1) +
  facet_wrap(~state, ncol = 1) +
  scale_fill_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "CoC-level target distribution over time, by state",
       x = "target_year", y = "homeless rate per 10k") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Graph 5 & 6: Predictor distributions (small multiples)

```{r graph5}
# pick the coc_/state_ numeric predictor columns (drop IDs, target, controls)
predictor_cols <- df %>%
  select(-state, -state_abbr, -coc_number, -coc_name, -predictor_year, -target_year,
         -target_homeless_rate_per_10k, -control_state_florida, -control_time_index) %>%
  names()

df_long <- df %>%
  select(all_of(predictor_cols)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# page 1: first half of predictors
page1_vars <- predictor_cols[1:20]

df_long %>%
  filter(variable %in% page1_vars) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free", ncol = 4) +
  labs(title = "Predictor distributions (page 1/2)", x = NULL, y = "count") +
  theme_minimal(base_size = 8)
```

```{r graph6}
# page 2: remaining predictors
page2_vars <- predictor_cols[21:length(predictor_cols)]

df_long %>%
  filter(variable %in% page2_vars) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free", ncol = 4) +
  labs(title = "Predictor distributions (page 2/2)", x = NULL, y = "count") +
  theme_minimal(base_size = 8)
```

## Graph 7 & 8: Predictor trends over time by state

```{r graph7}
state_predictor_trends <- df %>%
  select(state, predictor_year, all_of(predictor_cols)) %>%
  pivot_longer(all_of(predictor_cols), names_to = "variable", values_to = "value") %>%
  group_by(state, predictor_year, variable) %>%
  summarise(state_mean = mean(value, na.rm = TRUE), .groups = "drop")

state_predictor_trends %>%
  filter(variable %in% page1_vars) %>%
  ggplot(aes(x = predictor_year, y = state_mean, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 4) +
  scale_color_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "Predictor trends over time by state (page 1/2)",
       x = "predictor_year", y = "state mean") +
  theme_minimal(base_size = 8)
```

```{r graph8}
state_predictor_trends %>%
  filter(variable %in% page2_vars) %>%
  ggplot(aes(x = predictor_year, y = state_mean, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 4) +
  scale_color_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "Predictor trends over time by state (page 2/2)",
       x = "predictor_year", y = "state mean") +
  theme_minimal(base_size = 8)
```

## Graph 9: Clustered predictor correlation heatmap

```{r graph9}
cor_matrix <- df %>%
  select(all_of(predictor_cols)) %>%
  cor(use = "pairwise.complete.obs")

# average-linkage clustering to reorder variables
hc <- hclust(as.dist(1 - cor_matrix), method = "average")
ordered_vars <- rownames(cor_matrix)[hc$order]

cor_long <- melt(cor_matrix) %>%
  mutate(Var1 = factor(Var1, levels = ordered_vars),
         Var2 = factor(Var2, levels = ordered_vars))

ggplot(cor_long, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                        midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Clustered predictor correlation heatmap",
       subtitle = "Pairwise-complete Pearson r; average-linkage clustered order",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Graph 10: Target vs key predictors

```{r graph10}
key_predictors <- c(
  "coc_domestic_migration_rate_per_1000", "coc_hic_psh_beds_per_10k",
  "coc_hic_temporary_beds_per_10k", "coc_housing_cost_burdened_households_pct",
  "coc_housing_units_per_1000_residents", "coc_permits_per_1000_housing_units",
  "coc_population_density_per_sq_mile", "coc_poverty_all_pct",
  "coc_real_median_household_income_2025_usd", "coc_unemployment_rate_pct",
  "state_real_median_rent_2025_usd", "state_rental_vacancy_rate"
)

df %>%
  select(state, predictor_year, target_homeless_rate_per_10k, all_of(key_predictors)) %>%
  pivot_longer(all_of(key_predictors), names_to = "predictor", values_to = "value") %>%
  ggplot(aes(x = value, y = target_homeless_rate_per_10k, color = predictor_year)) +
  geom_point(alpha = 0.5, size = 1) +
  facet_grid(state ~ predictor, scales = "free_x") +
  scale_color_viridis_c(name = "year") +
  labs(title = "Target vs key predictors (no fitted line; descriptive)",
       subtitle = "Faceted by state (rows) x predictor (cols); colour = predictor_year",
       x = NULL, y = "homeless rate per 10k") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
gap_trends <- df %>%
  group_by(state, target_year) %>%
  summarise(mean_rate = mean(target_homeless_rate_per_10k), .groups = "drop") %>%
  pivot_wider(names_from = state, values_from = mean_rate) %>%
  mutate(gap = California - Florida)

ggplot(gap_trends, aes(x = target_year, y = gap)) +
  geom_line(color = "darkred", linewidth = 1.2) +
  geom_point(color = "darkred", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
 labs(
  title = "CA - FL gap in mean homeless rate over time",
  subtitle = "Unweighted mean across CoCs; positive = CA higher than FL",
  x = "Target year",
  y = "Gap in homeless rate per 10k (CA - FL)"
) +
theme_minimal(base_size = 15) +
theme(
  plot.title = element_text(size = 19, face = "bold"),
  plot.subtitle = element_text(size = 15),
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 14)
)
```

```{r}
library(sf)
library(dplyr)
library(purrr)

# 1. List every .shp file in each folder
ca_files <- list.files("~/Downloads/California", pattern = "\\.shp$", full.names = TRUE)
fl_files <- list.files("~/Downloads/Florida", pattern = "\\.shp$", full.names = TRUE)

# 2. Read every shapefile in, pulling the CoC number straight from the filename
#    (CA_510.shp -> "CA_510")
ca_boundaries <- map_dfr(ca_files, function(f) {
  st_read(f, quiet = TRUE) %>%
    mutate(coc_number_raw = tools::file_path_sans_ext(basename(f)))
})

fl_boundaries <- map_dfr(fl_files, function(f) {
  st_read(f, quiet = TRUE) %>%
    mutate(coc_number_raw = tools::file_path_sans_ext(basename(f)))
})

# 3. Combine both states
coc_boundaries <- bind_rows(ca_boundaries, fl_boundaries)

# 4. Convert underscore format (CA_510) to your data's hyphen format (CA-510)
coc_boundaries <- coc_boundaries %>%
  mutate(coc_number = gsub("_", "-", coc_number_raw))

# 5. Sanity check
nrow(coc_boundaries)
head(coc_boundaries$coc_number)

# 6. Pull the two years you want to map
map_years <- data_all %>%
  filter(target_year %in% c(2012, 2025)) %>%
  select(coc_number, target_year, target_homeless_rate_per_10k, state)

# 7. Join rate data onto the boundary polygons
map_data <- coc_boundaries %>%
  inner_join(map_years, by = "coc_number")

nrow(map_data)
```
# sanity check -- did the join
```{r}
library(car)        # for Anova()
library(tidyverse)   # dplyr, ggplot2

# Step 1: Read in the data ---------------------------------------------------
dat <- read_excel("CA_FL_LASSO_MODEL_INPUT_v2.xlsx", sheet = "LASSO Model Data")

# Step 2: Drop redundant/duplicate ID columns ---------------------------------
# These caused the 76 "singularity" NAs in the earlier fit3 -- each one is just
# a duplicate label for information already captured elsewhere:
#   coc_name         -> same CoC as coc_number, just spelled out
#   state_abbr        -> same state as `state`, just abbreviated
#   target_year       -> always predictor_year + 1, so it's redundant with predictor_year
#   control_time_index / control_state_florida -> re-encodings of predictor_year / state
dat_model <- dat %>%
  select( -coc_number,
    -coc_name,
    -state_abbr,
    -target_year,
    -control_time_index,
    -control_state_florida)

# Step 3: Log-transform the right-skewed predictors ---------------------------
# From the EDA: these coc_-level variables had skewness > 1 (piled low, long
# right tail). Use log(x) when every value is positive and log1p(x) when a
# variable contains genuine zeros. This matches the transformations used by
# the LASSO model below.
data_log <- dat_model

log_vars <- c(
  "coc_population_density_per_sq_mile",
  "coc_real_gdp_per_capita_2017_usd",
  "coc_hic_psh_beds_per_10k",
  "coc_hic_temporary_beds_per_10k",
  "coc_permits_per_1000_housing_units",
  "coc_permits_value_per_1000_housing_units_2025_usd",
  "coc_real_median_household_income_2025_usd",
  "coc_real_per_capita_personal_income_2025_usd",
  "coc_relative_home_price_index_2000_base"
)

# Use log(x) when every value is positive.
# Use log(1 + x) when the variable contains zeros.
for (v in log_vars) {
  if (min(data_log[[v]]) > 0) {
    data_log[[v]] <- log(data_log[[v]])
  } else {
    data_log[[v]] <- log1p(data_log[[v]])
  }
}

# The target is always positive, so use ordinary log()
data_log$target_homeless_rate_per_10k <-
  log(data_log$target_homeless_rate_per_10k)

# Fit the multiple regression using the transformed data
fit_log <- lm(
  target_homeless_rate_per_10k ~ .,
  data = data_log
)

coef_matrix <- summary(fit_log)$coefficients

coef_table <- data.frame(
  Predictor = rownames(coef_matrix),
  Estimate = coef_matrix[, 1],
  Std_Error = coef_matrix[, 2],
  t_value = coef_matrix[, 3],
  p_value = coef_matrix[, 4],
  row.names = NULL
)

coef_table
summary(fit_log)$r.squared
summary(fit_log)$adj.r.squared
summary(fit_log)$sigma
summary(fit_log)$fstatistic

# ---------------------------------------------------------------------------
# Reading the summary() output:
#   Estimate    -- change in log(Y) per 1-unit increase in X (or per log-unit
#                  increase in X, for the predictors we also logged above)
#   Std. Error  -- uncertainty around that estimate
#   t value / Pr(>|t|) -- test of whether that coefficient is significantly
#                          different from 0 (p < .05 = "significant")
#   R-squared   -- proportion of variance in log(Y) explained by the model
#   F-statistic -- tests whether the model as a WHOLE explains more than
#                  just guessing the mean every time
# ---------------------------------------------------------------------------

# Step 6: Confidence intervals for the coefficients -----------------------------
confint(fit_log)   # 95% CI for each beta -- if it crosses 0, not significant

# Step 7: F-test for each individual predictor, holding others fixed -----------
Anova(fit_log)

# Step 8: Model diagnostics on the log scale -------------------------------------
plot(fit_log, 1)   # residuals vs. fitted -- want a random scatter, no funnel shape
plot(fit_log, 2)   # base R's built-in Q-Q version
```
# LASSO

## Why LASSO and not ordinary regression

```{r lasso_setup}
library(glmnet)
library(coefplot)

# Define every object used below so the LASSO section runs in a clean session.
data_all <- dat
target <- "target_homeless_rate_per_10k"
controls <- c("control_state_florida", "control_time_index")

excluded_columns <- c(
  "state", "state_abbr", "coc_number", "coc_name",
  "predictor_year", "target_year", target, controls
)

predictors <- setdiff(names(data_all), excluded_columns)

length(predictors)   # should return 38
```

We have 38 predictors, and the EDA found **18 pairs correlated at |r| >= 0.80**.
With that much overlap, ordinary least squares gives unstable coefficients --
small changes in the data swing the estimates around.

LASSO adds a penalty that shrinks coefficients toward zero and sets the least
useful ones **exactly** to zero. The result is a smaller, steadier model.

## Log transformations

We log the target because of the right skew seen above. We also log nine
predictors that are dollar amounts, counts, or index values with long tails.

```{r}
# the nine predictors we are logging
log_vars <- c("coc_population_density_per_sq_mile",
              "coc_real_gdp_per_capita_2017_usd",
              "coc_hic_psh_beds_per_10k",
              "coc_hic_temporary_beds_per_10k",
              "coc_permits_per_1000_housing_units",
              "coc_permits_value_per_1000_housing_units_2025_usd",
              "coc_real_median_household_income_2025_usd",
              "coc_real_per_capita_personal_income_2025_usd",
              "coc_relative_home_price_index_2000_base")

# check the smallest value of each -- log() needs values above zero
sapply(data_all[, log_vars], min)
```

Two of them contain **real zeros**: 60 CoC-years have no permanent-supportive-housing
beds and one has no temporary beds. Those are genuine observations, not missing
data, but `log(0)` is undefined.

The fix is `log1p(x)`, which is just `log(1 + x)`. It turns 0 into 0 and leaves
every other value's ordering unchanged -- no row dropped, no made-up number added.

```{r}
# make a copy so the original data is untouched
data_log <- data_all

for (v in log_vars) {
  if (min(data_log[[v]]) > 0) {
    data_log[[v]] <- log(data_log[[v]])      # safe to log directly
  } else {
    data_log[[v]] <- log1p(data_log[[v]])    # has zeros -> use log(1 + x)
  }
}

# log the target too (its smallest value is 4.83, so plain log is fine)
data_log[[target]] <- log(data_log[[target]])
```

## Building the design matrix

```{r}
# keep only the columns that go into the model
fit_data <- data_log[, c(target, controls, predictors)]

# Y is the response, X is the matrix of predictors.
# model.matrix() builds X and handles any categorical variables for us.
# The [, -1] drops the leading column of 1s, which glmnet adds itself.
Y <- fit_data[[target]]
X <- model.matrix(target_homeless_rate_per_10k ~ ., data = fit_data)[, -1]

dim(X)     # 887 rows, 40 columns (38 predictors + 2 controls)
```

## Choosing lambda by cross validation

`lambda` controls how hard the penalty shrinks. Cross validation estimates the
prediction error at each lambda so we can pick one.

We also **force the two controls into the model** by giving them a penalty
factor of 0, so state and year are always adjusted for no matter how strong the
penalty gets.

```{r}
# penalty factor: 0 = never shrink this variable, 1 = shrink normally
pf <- ifelse(colnames(X) %in% controls, 0, 1)

set.seed(10)   # cross validation splits randomly, so we fix the seed
fit_cv <- cv.glmnet(X, Y, alpha = 1, nfolds = 10,
                    intercept = TRUE, penalty.factor = pf)
# alpha = 1 means LASSO (alpha = 0 would be Ridge)

fit_cv$lambda.min    # lambda with the lowest CV error
fit_cv$lambda.1se    # simplest lambda within one standard error of the best
```

```{r lasso_cv_plot, echo=FALSE, out.width="90%"}
# Write the base-R glmnet plot to a file, then insert that file. This avoids
# blank inline graphics devices in Positron.
plot_file <- "lasso_cv_plot.png"

png(
  filename = plot_file,
  width = 1200,
  height = 900,
  res = 150,
  bg = "white"
)

plot(fit_cv)
dev.off()

knitr::include_graphics(plot_file)
```

Reading this plot: red dots are the mean cross-validation error at each lambda,
and the bars are one standard error either side. On the displayed
$-\log(\lambda)$ axis, the left dashed line is `lambda.1se` and the right is
`lambda.min`. The numbers along the top count how
many coefficients are still non-zero.

```{r}
# how many variables does each choice keep?
coef_min <- extract.coef(fit_cv, lambda = "lambda.min")
coef_1se <- extract.coef(fit_cv, lambda = "lambda.1se")

nrow(coef_min) - 1   # minus 1 for the intercept -- expect 27
nrow(coef_1se) - 1   # expect 16
coef_min
```


```{r}
#| fig-height: 4.5
# the coefficient path: each line is one predictor shrinking toward zero
plot(glmnet(X, Y, alpha = 1, penalty.factor = pf), xvar = "lambda", label = TRUE)
```

## Relaxed LASSO

LASSO picks variables but does not give us p-values for its own choices. The
standard fix is to take the variables it picked and refit them with plain `lm()`.
This is called the **relaxed LASSO**.

```{r}
# the variable names LASSO selected at lambda.min
var_min <- setdiff(coef_min$Coefficient, "(Intercept)")

# refit those variables with ordinary regression
fit_lm <- lm(as.formula(paste(target, "~", paste(var_min, collapse = " + "))),
             data = fit_data)
summary(fit_lm)
summary(fit_lm)$adj.r.squared     # expect 0.6436
```

```{r}
# how many of the 27 are actually significant?
sum(summary(fit_lm)$coefficients[-1, 4] < 0.05)   # expect 18
```

Only 18 of 27 clear the 0.05 bar, so there is still trimming to do.

## Fine-tuning with backward selection

Backward selection: find the least significant substantive predictor, drop it,
refit, and repeat until every removable predictor left is significant. State
and year remain in the model because they were prespecified controls.

```{r}
fit_back <- fit_lm

repeat {
  # Anova() gives one p-value per predictor
  anova_tbl <- Anova(fit_back, type = "II")
  pvals <- anova_tbl[["Pr(>F)"]]
  names(pvals) <- rownames(anova_tbl)

  # Controls stay for adjustment even when their own p-values exceed 0.05.
  removable_pvals <- pvals[
    !is.na(pvals) &
      !names(pvals) %in% c("Residuals", controls)
  ]

  # Stop once every remaining non-control predictor is significant.
  if (length(removable_pvals) == 0 ||
      max(removable_pvals) <= 0.05) break

  worst <- names(removable_pvals)[which.max(removable_pvals)]
  cat("dropping", worst,
      " p =", round(max(removable_pvals), 3), "\n")

  fit_back <- update(fit_back, paste(". ~ . -", worst))
}

fit_final <- fit_back
```

```{r}
length(coef(fit_final)) - 1              # expect 21 predictors

summary(fit_final)
summary(fit_final)$adj.r.squared         # expect about 0.6445
```

Six substantive predictors came out and the adjusted R-squared barely moved
(0.6436 to about 0.6445). That is the point -- those six were costing us model
complexity without adding explanation.

Worth noticing: **`control_state_florida` remains in the model as a prespecified
control but is not individually significant.** After accounting for the
selected CoC-level variables, the model does not detect a remaining adjusted
state difference.

## Checking the model assumptions

```{r}
#| fig-height: 4.5
plot(fit_final, 1)     # residuals vs fitted -- we want a flat line, even scatter
```

```{r}
#| fig-height: 4.5
plot(fit_final, 2)     # Q-Q plot -- we want points close to the diagonal
```

The residual band is roughly constant across the range, which is exactly what
the log transform was for. On the original rate scale the spread would grow with
the prediction. The Q-Q plot tracks the diagonal through the middle with mild
tail departures -- not perfect, but reasonable.

## Reading the final model

```{r}
# sort the coefficients by how statistically strong they are
final_tbl <- summary(fit_final)$coefficients
final_tbl <- final_tbl[order(-abs(final_tbl[, 3])), ]
round(head(final_tbl, 10), 4)
```

Because the response is logged, a coefficient is roughly the **percent change**
in next-year homelessness for a one-unit rise in the predictor. For the six
predictors that are also logged, both sides are on a log scale, so the
coefficient is an **elasticity** -- a 1% rise in the predictor goes with a beta
percent change in the rate.

**Shelter capacity is the strongest term** (temporary beds, elasticity +0.358).
This does **not** mean shelters cause homelessness. Communities build capacity
where homelessness is already high, and sheltered counts are taken in those
beds -- the association runs from need to capacity.

## A limit worth stating

The p-values above are optimistic. LASSO chose the variables using the same 887
rows that `lm()` then tested them on, and backward selection used those rows
again. Treat them as a guide to which terms are strong, not as formal tests.

Also, `cv.glmnet` shuffles rows at **random**, so a model can be trained partly
on 2024 and evaluated on 2015. That is fine for picking lambda, but it means
nothing above tests whether we can predict a *future* year. Different years
from the same CoC can also appear in both training and validation folds, so the
cross-validation error may be optimistic for predicting entirely new CoCs.

---
# Neural Network

## What changes for a neural network

A neural network can capture curved and interacting relationships that a linear
model cannot. Four things differ from the textbook classification example:

| Textbook setup | What we do here | Why |
|---|---|---|
| Classification, softmax output | **Regression**: one linear output unit, MSE loss | Our target is a continuous rate |
| Random train/test split | **Rolling origin** over 8 future years | Random splits leak future information |
| Random internal validation split | Hold out the most recent training year | Same reason |
| No scaling | Scale using **training rows only** | The model must not see the validation year |

```{r}
#| eval: false
# Keras runs on Python. This loads the environment set up in Module 7.
library(reticulate)
use_condaenv("r", required = TRUE)
library(keras3)
```

```{r}
#| include: false
suppressMessages({
  library(reticulate)
  try(use_condaenv("r", required = TRUE), silent = TRUE)
  library(keras3)
})
```

## Setting up rolling-origin validation

Instead of one random split, we validate **eight times**. Each time the model
trains only on years before the validation year, then predicts that year once.

```{r}
# the eight years we validate on (2021 is absent -- COVID)
valid_years <- c(2017, 2018, 2019, 2020, 2022, 2023, 2024, 2025)

# the columns the network sees: 2 controls + 38 predictors = 40 inputs
nn_features <- c(controls, predictors)

# use the ORIGINAL (unlogged) data here, so results compare to the LASSO
X_all <- as.matrix(data_all[, nn_features])
y_all <- data_all[[target]]

# show how the training window grows
for (vy in valid_years) {
  cat("validate", vy, "-> train on", sum(data_all$target_year < vy),
      "rows, score", sum(data_all$target_year == vy), "rows\n")
}
```

## Building a network

```{r}
# A small helper so we can try different shapes without repeating ourselves.
# units = how many neurons in each hidden layer.
build_net <- function(n_inputs, units, dropout = 0) {
  model <- keras_model_sequential(input_shape = c(n_inputs))

  for (u in units) {
    model <- model %>% layer_dense(units = u, activation = "relu")
    if (dropout > 0) model <- model %>% layer_dropout(rate = dropout)
  }

  # ONE output unit with no activation -- this makes it a regression
  model <- model %>% layer_dense(units = 1)

  model %>% compile(optimizer = "rmsprop", loss = "mse", metrics = c("mae"))
  model
}
```

## Fitting one fold

This function does the whole job for a single validation year: scale on the
training rows, decide how many epochs to use, refit, and predict.

```{r}
fit_one_fold <- function(valid_year, units, dropout, seed) {

  train_rows <- which(data_all$target_year <  valid_year)
  valid_rows <- which(data_all$target_year == valid_year)

  # --- scaling, computed on TRAINING rows only ---
  mu <- colMeans(X_all[train_rows, ])
  sdv <- apply(X_all[train_rows, ], 2, sd)
  sdv[sdv == 0] <- 1                       # avoid dividing by zero

  x_train <- scale(X_all[train_rows, ], center = mu, scale = sdv)
  x_valid <- scale(X_all[valid_rows, ], center = mu, scale = sdv)

  # scale the target too, then undo it on the predictions later
  y_mu <- mean(y_all[train_rows]); y_sd <- sd(y_all[train_rows])
  y_train <- (y_all[train_rows] - y_mu) / y_sd

  # --- inner split: hold out the most recent training year ---
  # (a random split here would let future years influence when we stop)
  yrs <- data_all$target_year[train_rows]
  inner_train <- which(yrs <  max(yrs))
  inner_valid <- which(yrs == max(yrs))

  set_random_seed(seed)

  # fit while watching the held-out year, and stop when it stops improving
  m0 <- build_net(ncol(x_train), units, dropout)
  history <- m0 %>% fit(
    x_train[inner_train, ], y_train[inner_train],
    validation_data = list(x_train[inner_valid, ], y_train[inner_valid]),
    epochs = 300, batch_size = 32, verbose = 0,
    callbacks = list(callback_early_stopping(monitor = "val_loss",
                                             patience = 20,
                                             restore_best_weights = TRUE)))

  best_epochs <- which.min(history$metrics$val_loss)

  # --- refit on the full training window for that many epochs ---
  set_random_seed(seed)
  m <- build_net(ncol(x_train), units, dropout)
  m %>% fit(x_train, y_train, epochs = best_epochs,
            batch_size = 32, verbose = 0)

  # predict, then undo the target scaling
  pred <- as.numeric(predict(m, x_valid, verbose = 0)) * y_sd + y_mu

  data.frame(valid_year = valid_year,
             actual = y_all[valid_rows],
             predicted = pred)
}
```

## Running all eight folds

Network training is random, so on a panel this small a single run is mostly
noise. We fit each fold under **five different seeds** and average the
predictions.

```{r}
#| eval: false
# This takes roughly 8 minutes for one architecture (8 folds x 5 seeds,
# about 11 seconds each). Set eval: true to actually run it.
seeds <- c(10, 42, 123, 2024, 31337)

run_network <- function(units, dropout) {
  out <- list()
  for (vy in valid_years) {
    # fit the fold once per seed
    folds <- lapply(seeds, function(s) fit_one_fold(vy, units, dropout, s))

    # the actual values are identical every time, so keep the first result
    # and replace its predictions with the average across all five seeds
    result <- folds[[1]]
    result$predicted <- rowMeans(sapply(folds, function(f) f$predicted))

    out[[length(out) + 1]] <- result
  }
  do.call(rbind, out)
}

nn_16_8   <- run_network(units = c(16, 8), dropout = 0)     # standard shape
nn_linear <- run_network(units = c(),      dropout = 0)     # no hidden layers
```

## Scoring the results

We use the same three measures throughout the project, so every model can be
compared on equal terms.

```{r}
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae  <- function(actual, pred) mean(abs(actual - pred))
r2   <- function(actual, pred) 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)

score <- function(df, label) {
  data.frame(model = label,
             n     = nrow(df),
             RMSE  = rmse(df$actual, df$predicted),
             MAE   = mae(df$actual, df$predicted),
             R2    = r2(df$actual, df$predicted))
}
```

```{r}
#| eval: false
rbind(score(nn_16_8,   "neural net 16-8"),
      score(nn_linear, "no hidden layers (= OLS)"))
```

Running the full set of five architectures (in `fit_neural_net.R`) gives:

| Model | RMSE | MAE | R² |
|---|---:|---:|---:|
| Neural net, 32-16 with dropout | **12.64** | 8.43 | **0.731** |
| Neural net, 16-8 | 12.92 | 9.11 | 0.719 |
| LASSO with state interactions | 14.69 | 10.14 | 0.636 |
| Neural net, no hidden layers (= OLS) | 16.06 | 11.46 | 0.565 |
| State + time baseline | 21.93 | 14.98 | 0.190 |

Two things to read off this table.

**The networks beat the LASSO.** The best is a 14% reduction in RMSE.

**The no-hidden-layer version tells us why.** With no hidden layers the network
*is* ordinary least squares, and it does **worse** than the LASSO (16.06 vs
14.69) -- exactly as expected, since that is the situation the LASSO penalty is
built for. So the network's advantage comes from **nonlinearity**, not from
anything else in how it was set up.

## The catch

```{r}
#| eval: false
# look at how much the answer moves between seeds for a single fold
sapply(seeds, function(s) {
  f <- fit_one_fold(2025, units = c(16, 8), dropout = 0, seed = s)
  rmse(f$actual, f$predicted)
})
```

Across the whole run, a single 16-8 network scores anywhere from **14.38 to
16.29** depending only on the random seed. The five-seed average scores 12.92 --
better than *any* individual seed. So a large part of the improvement is
averaging away instability, not one network being better.

An unlucky seed loses to the LASSO outright. The LASSO, by contrast, is
deterministic: same data, same answer, every time.

The honest statement is: **an ensemble of five networks beats the LASSO; a
single network is better on average but unreliable.**

## The result that matters most

We also asked whether either model finds anything beyond simply knowing **last
year's rate** for that CoC.

| Model | RMSE | R² |
|---|---:|---:|
| LASSO: state + time + prior rate | **8.89** | 0.866 |
| Neural net on prior rate alone | 8.93 | 0.865 |
| Prior rate only | 9.12 | 0.859 |
| Neural net: prior rate + 38 predictors | 10.05 | 0.829 |
| LASSO: prior rate + 38 predictors | 8.89 | 0.866 |

Given the prior year's rate, **the LASSO shrank all 38 predictors to exactly
zero** -- in every fold, under both lambda rules. And **adding those 38
predictors made the neural network worse**, from 8.93 up to 10.05.

The reason is the difference between the two methods. LASSO can say "ignore
these" and pay nothing for having been offered them. A neural network cannot --
every input stays wired into the first layer, so 38 uninformative predictors
inject noise it has no way to discard.

Two model families with completely different failure modes reached the same
conclusion: **next-year homelessness is dominated by its own prior level, and
the predictor set adds little on top of that.**

---

# Summary {-}

```{r}
#| echo: false
data.frame(
  Stage = c("Data", "LASSO", "LASSO", "Neural net", "Both"),
  Result = c("887 CoC-years, 70 CoCs, 38 predictors, 0 missing values",
             "27 variables at lambda.min, 16 at lambda.1se",
             "Final model: 20 predictors, adjusted R-squared 0.645",
             "Best out-of-time RMSE 12.64, R-squared 0.731",
             "Predictors add nothing once last year's rate is known"))
```

Everything above describes **association, not causation**. Two states observed
over a short, strongly time-ordered panel cannot support causal claims, and the
shelter-bed result is the clearest reminder of why: it is one of the strongest
predictors in every model, and it almost certainly reflects capacity following
need rather than the other way round.

# Reproduce {-}

```{r}
#| eval: false
# the full production versions of these three stages
source("scripts/build_expanded_lasso_input_v2.R")   # data
source("scripts/fit_lasso_logged.R")                # LASSO
source("scripts/fit_neural_net.R")                  # neural network (~35 min)
```