---
title: "Rolling-Origin LASSO"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Purpose

This model predicts future CoC homelessness rates in chronological order. For
each test year, it:

1. fits candidate LASSO penalties using earlier years;
2. selects lambda using the most recent available training year;
3. refits the model using all observations before the test year; and
4. evaluates predictions on the untouched test year.

## Packages and data

Keep this Rmd and `CA_FL_LASSO_MODEL_INPUT_v2.xlsx` in the same folder.

```{r packages-and-data}
library(readxl)
library(dplyr)
library(glmnet)
library(ggplot2)
library(knitr)

input_path <- "CA_FL_LASSO_MODEL_INPUT_v2.xlsx"

if (!file.exists(input_path)) {
  stop(
    "Cannot find ", input_path,
    ". Put the Excel workbook in the same folder as this Rmd."
  )
}

data_all <- read_excel(
  input_path,
  sheet = "LASSO Model Data"
)

target <- "target_homeless_rate_per_10k"
controls <- c(
  "control_state_florida",
  "control_time_index"
)

required_columns <- c(
  "state",
  "coc_number",
  "coc_name",
  "target_year",
  target,
  controls
)

missing_columns <- setdiff(
  required_columns,
  names(data_all)
)

if (length(missing_columns) > 0) {
  stop(
    "The workbook is missing required columns: ",
    paste(missing_columns, collapse = ", ")
  )
}
```

## Choose predictors

Names, identifiers, and year columns are not used as penalized predictors.
State and time enter separately as unpenalized controls.

```{r choose-predictors}
excluded_columns <- c(
  "state",
  "state_abbr",
  "coc_number",
  "coc_name",
  "predictor_year",
  "target_year",
  target,
  controls
)

predictors <- setdiff(
  names(data_all),
  excluded_columns
)

length(predictors)
```

## Log transformations

The response and nine right-skewed predictors are transformed in the same way
as in the main analysis. Ordinary `log(x)` is used when every value is positive;
`log1p(x)` is used when genuine zeros occur.

```{r log-transformations}
log_vars <- c(
  "coc_population_density_per_sq_mile",
  "coc_real_gdp_per_capita_2017_usd",
  "coc_hic_psh_beds_per_10k",
  "coc_hic_temporary_beds_per_10k",
  "coc_permits_per_1000_housing_units",
  "coc_permits_value_per_1000_housing_units_2025_usd",
  "coc_real_median_household_income_2025_usd",
  "coc_real_per_capita_personal_income_2025_usd",
  "coc_relative_home_price_index_2000_base"
)

missing_log_vars <- setdiff(
  log_vars,
  names(data_all)
)

if (length(missing_log_vars) > 0) {
  stop(
    "The workbook is missing log-transformed predictors: ",
    paste(missing_log_vars, collapse = ", ")
  )
}

data_log <- data_all

for (v in log_vars) {
  if (min(data_log[[v]], na.rm = TRUE) > 0) {
    data_log[[v]] <- log(data_log[[v]])
  } else {
    data_log[[v]] <- log1p(data_log[[v]])
  }
}

if (min(data_log[[target]], na.rm = TRUE) <= 0) {
  stop("The target must be positive before applying log().")
}

data_log[[target]] <- log(
  data_log[[target]]
)
```

## Build the LASSO inputs

`Y` is the logged response. `X` is the numerical predictor matrix. The two
controls receive penalty factors of zero, meaning they remain in every model.

```{r build-inputs}
fit_data <- data_log[
  ,
  c(target, controls, predictors)
]

if (anyNA(fit_data)) {
  stop(
    "The modeling columns contain missing values. ",
    "Resolve them before fitting the rolling LASSO."
  )
}

Y <- fit_data[[target]]

X <- model.matrix(
  target_homeless_rate_per_10k ~ .,
  data = fit_data
)[, -1, drop = FALSE]

pf <- ifelse(
  colnames(X) %in% controls,
  0,
  1
)

dim(X)
table(pf)
```

## Run the rolling-origin LASSO

For example, when predicting 2025, candidate lambdas are fitted through 2023,
lambda is selected using 2024, the model is refitted through 2024, and the
final predictions are evaluated on 2025.

```{r rolling-origin-lasso}
rolling_years <- c(
  2017, 2018, 2019, 2020,
  2022, 2023, 2024, 2025
)

missing_test_years <- setdiff(
  rolling_years,
  unique(data_all$target_year)
)

if (length(missing_test_years) > 0) {
  stop(
    "No observations were found for test years: ",
    paste(missing_test_years, collapse = ", ")
  )
}

rolling_results <- vector(
  "list",
  length(rolling_years)
)

rolling_importance <- vector(
  "list",
  length(rolling_years)
)

for (i in seq_along(rolling_years)) {

  test_year <- rolling_years[i]

  # Outer split: the test year remains completely untouched.
  outer_train <- which(
    data_all$target_year < test_year
  )

  outer_test <- which(
    data_all$target_year == test_year
  )

  # Inner split: use the latest training year to select lambda.
  years_in_training <- data_all$target_year[
    outer_train
  ]

  tune_year <- max(
    years_in_training
  )

  inner_train <- outer_train[
    years_in_training < tune_year
  ]

  inner_valid <- outer_train[
    years_in_training == tune_year
  ]

  if (length(inner_train) == 0 ||
      length(inner_valid) == 0 ||
      length(outer_test) == 0) {
    stop(
      "Insufficient chronological data for test year ",
      test_year
    )
  }

  # Fit the complete LASSO path using years before the tuning year.
  tuning_fit <- glmnet(
    x = X[inner_train, , drop = FALSE],
    y = Y[inner_train],
    alpha = 1,
    penalty.factor = pf
  )

  # Predict the tuning year at every candidate lambda.
  tuning_predictions <- predict(
    tuning_fit,
    newx = X[inner_valid, , drop = FALSE],
    s = tuning_fit$lambda
  )

  # Calculate tuning-year MSE for every lambda.
  tuning_mse <- colMeans(
    sweep(
      tuning_predictions,
      MARGIN = 1,
      STATS = Y[inner_valid],
      FUN = "-"
    )^2
  )

  # Choose the lambda with the lowest next-year validation MSE.
  best_lambda <- tuning_fit$lambda[
    which.min(tuning_mse)
  ]

  # Refit using every observation before the final test year.
  final_rolling_fit <- glmnet(
    x = X[outer_train, , drop = FALSE],
    y = Y[outer_train],
    alpha = 1,
    lambda = tuning_fit$lambda,
    penalty.factor = pf
  )

  # Predict the untouched test year.
  predicted_log_rate <- predict(
    final_rolling_fit,
    newx = X[outer_test, , drop = FALSE],
    s = best_lambda
  )

  # Return predictions to homelessness rates per 10,000.
  predicted_rate <- exp(
    as.numeric(predicted_log_rate)
  )

  fold_coefficients <- as.vector(
    coef(
      final_rolling_fit,
      s = best_lambda
    )
  )[-1]

  number_nonzero <- sum(
    abs(fold_coefficients) > 1e-8
  )

  # Standardize coefficient magnitudes for comparison across units.
  training_sd <- apply(
    X[outer_train, , drop = FALSE],
    2,
    sd
  )

  rolling_importance[[i]] <- data.frame(
    test_year = test_year,
    predictor = colnames(X),
    coefficient = fold_coefficients,
    standardized_effect =
      fold_coefficients * training_sd,
    selected =
      abs(fold_coefficients) > 1e-8
  )

  # Save one row per CoC prediction.
  rolling_results[[i]] <- data.frame(
    state = data_all$state[outer_test],
    coc_number = data_all$coc_number[outer_test],
    coc_name = data_all$coc_name[outer_test],
    test_year = test_year,
    tune_year = tune_year,
    actual = data_all[[target]][outer_test],
    predicted = predicted_rate,
    lambda = best_lambda,
    nonzero_coefficients = number_nonzero
  )
}

rolling_lasso <- bind_rows(
  rolling_results
)

rolling_importance_all <- bind_rows(
  rolling_importance
)

head(rolling_lasso)
```

## Overall out-of-time performance

RMSE and MAE are expressed as homeless people per 10,000 residents. Predictive
R-squared compares the out-of-time predictions with predicting the pooled mean.

```{r overall-results}
rolling_lasso_score <- data.frame(
  observations = nrow(rolling_lasso),

  RMSE = sqrt(
    mean(
      (rolling_lasso$actual -
         rolling_lasso$predicted)^2
    )
  ),

  MAE = mean(
    abs(
      rolling_lasso$actual -
        rolling_lasso$predicted
    )
  ),

  R2 = 1 -
    sum(
      (rolling_lasso$actual -
         rolling_lasso$predicted)^2
    ) /
    sum(
      (rolling_lasso$actual -
         mean(rolling_lasso$actual))^2
    )
)

kable(
  round(rolling_lasso_score, 3),
  caption = "Pooled rolling-origin LASSO performance"
)
```

## Performance by test year

```{r results-by-year}
rolling_lasso_by_year <- rolling_lasso %>%
  group_by(test_year, tune_year) %>%
  summarise(
    observations = n(),
    RMSE = sqrt(
      mean((actual - predicted)^2)
    ),
    MAE = mean(
      abs(actual - predicted)
    ),
    lambda = first(lambda),
    nonzero_coefficients =
      first(nonzero_coefficients),
    .groups = "drop"
  )

kable(
  rolling_lasso_by_year,
  digits = 3,
  caption = "Rolling-origin performance by test year"
)
```

## Most consistently useful predictors

`folds_selected` shows how many of the eight chronological models selected a
predictor. `mean_abs_standardized_effect` measures its average predictive
magnitude when selected, after accounting for differences in predictor units.

```{r predictor-importance}
importance_table <- rolling_importance_all %>%
  filter(
    !predictor %in% controls
  ) %>%
  group_by(predictor) %>%
  summarise(
    folds_selected = sum(selected),
    selection_rate = mean(selected),

    mean_abs_standardized_effect =
      if (any(selected)) {
        mean(
          abs(standardized_effect[selected])
        )
      } else {
        0
      },

    mean_standardized_effect =
      if (any(selected)) {
        mean(
          standardized_effect[selected]
        )
      } else {
        0
      },

    .groups = "drop"
  ) %>%
  mutate(
    average_direction = case_when(
      mean_standardized_effect > 0 ~ "Positive",
      mean_standardized_effect < 0 ~ "Negative",
      TRUE ~ "None"
    )
  ) %>%
  select(
    -mean_standardized_effect
  ) %>%
  arrange(
    desc(folds_selected),
    desc(mean_abs_standardized_effect)
  )

kable(
  head(importance_table, 15),
  digits = 3,
  caption = "Top 15 rolling-origin LASSO predictors"
)
```

## Actual versus predicted values

```{r prediction-plot, fig.width=8, fig.height=6}
ggplot(
  rolling_lasso,
  aes(x = actual, y = predicted, color = state)
) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "gray45"
  ) +
  geom_point(
    alpha = 0.65,
    size = 1.8
  ) +
  scale_color_manual(
    values = c(
      "California" = "darkgreen",
      "Florida" = "purple"
    )
  ) +
  labs(
    title = "Rolling-origin LASSO: actual versus predicted rates",
    subtitle = "Dashed line represents perfect predictions",
    x = "Actual homelessness rate per 10,000",
    y = "Predicted homelessness rate per 10,000",
    color = "State"
  ) +
  theme_minimal(base_size = 14)
```

## Interpretation limits

- Each test year is predicted only from earlier observations.
- Lambda is selected chronologically and never uses the final test year.
- Selection frequency and standardized effect describe predictive usefulness,
  not statistical significance.
- Positive or negative coefficients describe associations, not causation.
- Observations from the same CoCs occur across years, so this tests prediction
  of future years for the studied CoCs rather than prediction for entirely new
  CoCs.


