---
title: "CA vs FL Homelessness EDA Graphs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages and data

```{r load}
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)

df <- read_excel(
  "outputs/lasso_model/CA_FL_LASSO_MODEL_INPUT_v2.xlsx",
  sheet = "LASSO Model Data"
)
```

## Graph 1: Raw target distribution

```{r graph1}
skew <- function(x) {
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  sum((x - m)^3) / n / s^3
}

skew_raw <- round(skew(df$target_homeless_rate_per_10k), 2)

ggplot(df, aes(x = target_homeless_rate_per_10k)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Target distribution: next-year homeless rate per 10k",
       subtitle = paste0("n=", nrow(df), ", skewness=", skew_raw),
       x = "target_homeless_rate_per_10k", y = "density") +
  theme_minimal()
```

## Graph 2: Log target distribution

```{r graph2}
skew_log <- round(skew(log(df$target_homeless_rate_per_10k)), 2)

ggplot(df, aes(x = log(target_homeless_rate_per_10k))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Log target distribution",
       subtitle = paste0("skewness=", skew_log),
       x = "log(target_homeless_rate_per_10k)", y = "density") +
  theme_minimal()
```

## Graph 3: CA vs FL target trends over time

```{r graph3}
state_trends <- df %>%
  group_by(state, target_year) %>%
  summarise(mean_rate = mean(target_homeless_rate_per_10k), .groups = "drop")

ggplot(df, aes(x = target_year, y = target_homeless_rate_per_10k, color = state)) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 1) +
  geom_line(data = state_trends, aes(x = target_year, y = mean_rate, color = state),
            linewidth = 1.2) +
  geom_point(data = state_trends, aes(x = target_year, y = mean_rate, color = state),
             size = 2.5) +
  scale_color_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "CA vs FL: next-year homeless rate over time",
       x = "target_year", y = "homeless rate per 10k") +
theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(size = 19, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 14)
  )
```

## Graph 4: CoC-level target boxplots by state, over time

```{r graph4}
ggplot(df, aes(x = factor(target_year), y = target_homeless_rate_per_10k, fill = state)) +
  geom_boxplot(outlier.size = 1) +
  facet_wrap(~state, ncol = 1) +
  scale_fill_manual(values = c("California" = "darkgreen", "Florida" = "purple")) +
  labs(title = "CoC-level target distribution over time, by state",
       x = "target_year", y = "homeless rate per 10k") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Graph 5 & 6: Predictor distributions (small multiples)

```{r graph5}
# pick the coc_/state_ numeric predictor columns (drop IDs, target, controls)
predictor_cols <- df %>%
  select(-state, -state_abbr, -coc_number, -coc_name, -predictor_year, -target_year,
         -target_homeless_rate_per_10k, -control_state_florida, -control_time_index) %>%
  names()

df_long <- df %>%
  select(all_of(predictor_cols)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# page 1: first half of predictors
page1_vars <- predictor_cols[1:20]

df_long %>%
  filter(variable %in% page1_vars) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free", ncol = 4) +
  labs(title = "Predictor distributions (page 1/2)", x = NULL, y = "count") +
  theme_minimal(base_size = 8)
```

```{r graph6}
# page 2: remaining predictors
page2_vars <- predictor_cols[21:length(predictor_cols)]

df_long %>%
  filter(variable %in% page2_vars) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free", ncol = 4) +
  labs(title = "Predictor distributions (page 2/2)", x = NULL, y = "count") +
  theme_minimal(base_size = 8)
```

## Graph 7 & 8: Predictor trends over time by state

```{r graph7}
state_predictor_trends <- df %>%
  select(state, predictor_year, all_of(predictor_cols)) %>%
  pivot_longer(all_of(predictor_cols), names_to = "variable", values_to = "value") %>%
  group_by(state, predictor_year, variable) %>%
  summarise(state_mean = mean(value, na.rm = TRUE), .groups = "drop")

state_predictor_trends %>%
  filter(variable %in% page1_vars) %>%
  ggplot(aes(x = predictor_year, y = state_mean, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 4) +
  scale_color_manual(values = c("California" = "black", "Florida" = "darkblue")) +
  labs(title = "Predictor trends over time by state (page 1/2)",
       x = "predictor_year", y = "state mean") +
  theme_minimal(base_size = 8)
```

```{r graph8}
state_predictor_trends %>%
  filter(variable %in% page2_vars) %>%
  ggplot(aes(x = predictor_year, y = state_mean, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 4) +
  scale_color_manual(values = c("California" = "black", "Florida" = "darkblue")) +
  labs(title = "Predictor trends over time by state (page 2/2)",
       x = "predictor_year", y = "state mean") +
  theme_minimal(base_size = 8)
```

## Graph 9: Clustered predictor correlation heatmap

```{r graph9}
cor_matrix <- df %>%
  select(all_of(predictor_cols)) %>%
  cor(use = "pairwise.complete.obs")

# average-linkage clustering to reorder variables
hc <- hclust(as.dist(1 - cor_matrix), method = "average")
ordered_vars <- rownames(cor_matrix)[hc$order]

cor_long <- melt(cor_matrix) %>%
  mutate(Var1 = factor(Var1, levels = ordered_vars),
         Var2 = factor(Var2, levels = ordered_vars))

ggplot(cor_long, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                        midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Clustered predictor correlation heatmap",
       subtitle = "Pairwise-complete Pearson r; average-linkage clustered order",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Graph 10: Target vs key predictors

```{r graph10}
key_predictors <- c(
  "coc_domestic_migration_rate_per_1000", "coc_hic_psh_beds_per_10k",
  "coc_hic_temporary_beds_per_10k", "coc_housing_cost_burdened_households_pct",
  "coc_housing_units_per_1000_residents", "coc_permits_per_1000_housing_units",
  "coc_population_density_per_sq_mile", "coc_poverty_all_pct",
  "coc_real_median_household_income_2025_usd", "coc_unemployment_rate_pct",
  "state_real_median_rent_2025_usd", "state_rental_vacancy_rate"
)

df %>%
  select(state, predictor_year, target_homeless_rate_per_10k, all_of(key_predictors)) %>%
  pivot_longer(all_of(key_predictors), names_to = "predictor", values_to = "value") %>%
  ggplot(aes(x = value, y = target_homeless_rate_per_10k, color = predictor_year)) +
  geom_point(alpha = 0.5, size = 1) +
  facet_grid(state ~ predictor, scales = "free_x") +
  scale_color_viridis_c(name = "year") +
  labs(title = "Target vs key predictors (no fitted line; descriptive)",
       subtitle = "Faceted by state (rows) x predictor (cols); colour = predictor_year",
       x = NULL, y = "homeless rate per 10k") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
gap_trends <- df %>%
  group_by(state, target_year) %>%
  summarise(mean_rate = mean(target_homeless_rate_per_10k), .groups = "drop") %>%
  pivot_wider(names_from = state, values_from = mean_rate) %>%
  mutate(gap = California - Florida)

ggplot(gap_trends, aes(x = target_year, y = gap)) +
  geom_line(color = "darkred", linewidth = 1.2) +
  geom_point(color = "darkred", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
 labs(
  title = "CA - FL gap in mean homeless rate over time",
  subtitle = "Unweighted mean across CoCs; positive = CA higher than FL",
  x = "Target year",
  y = "Gap in homeless rate per 10k (CA - FL)"
) +
theme_minimal(base_size = 15) +
theme(
  plot.title = element_text(size = 19, face = "bold"),
  plot.subtitle = element_text(size = 15),
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 14)
)
```

```{r coc-boundaries, eval=FALSE}
# Optional map setup. It requires local HUD CoC shapefiles in
# ~/Downloads/California and ~/Downloads/Florida, and is disabled so a missing
# map download never blocks the EDA or LASSO sections below.
library(sf)
library(dplyr)
library(purrr)

# 1. Find every .shp file, including files inside each CoC subfolder
ca_files <- list.files(
  "~/Downloads/California",
  pattern = "\\.shp$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

fl_files <- list.files(
  "~/Downloads/Florida",
  pattern = "\\.shp$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

# Skip maps when the optional local boundary downloads are unavailable. This
# lets the EDA and LASSO sections run in a notebook without map files.
if (length(ca_files) == 0 || length(fl_files) == 0) {
  message(
    "Skipping optional CoC maps: no shapefiles found in ~/Downloads/California ",
    "and/or ~/Downloads/Florida."
  )
  coc_boundaries <- NULL
  map_source <- NULL
  map_data <- NULL
} else {
  # 2. Read each boundary and pull its CoC number from its filename
#    (CA_510.shp -> "CA_510")
read_boundary <- function(f) {
  st_read(f, quiet = TRUE) %>%
    mutate(coc_number_raw = tools::file_path_sans_ext(basename(f)))
}

ca_boundaries <- map_dfr(ca_files, read_boundary)
fl_boundaries <- map_dfr(fl_files, read_boundary)

# 3. Combine both states
coc_boundaries <- bind_rows(ca_boundaries, fl_boundaries)

# 4. Convert underscore format (CA_510) to your data's hyphen format (CA-510)
coc_boundaries <- coc_boundaries %>%
  mutate(coc_number = gsub("_", "-", coc_number_raw))

# 5. Sanity checks: expect 71 boundary files and 71 distinct IDs
nrow(coc_boundaries)
n_distinct(coc_boundaries$coc_number)
head(coc_boundaries$coc_number)

# 6. Load the model data here so this map chunk also works by itself
map_source <- readxl::read_excel(
  "outputs/lasso_model/CA_FL_LASSO_MODEL_INPUT_v2.xlsx",
  sheet = "LASSO Model Data"
)

# Pull the two years you want to map
map_years <- map_source %>%
  filter(target_year %in% c(2012, 2025)) %>%
  select(coc_number, target_year, target_homeless_rate_per_10k, state)

# 7. Join rate data onto the boundary polygons
map_data <- coc_boundaries %>%
  inner_join(map_years, by = "coc_number")

nrow(map_data)
}
```
## CoC homelessness-rate maps

The states are plotted separately so each pair of maps fills the available
space. Both states use the same color scale, so their colors remain comparable.

```{r coc-maps-side-by-side, eval=FALSE, fig.width=16, fig.height=8.5, fig.retina=2, out.width='100%'}
if (is.null(map_data)) {
  message("Skipping optional static maps because CoC boundary files are unavailable.")
} else {
rate_limits <- range(
  map_data$target_homeless_rate_per_10k,
  na.rm = TRUE
)

make_state_map <- function(state_name) {
  map_data %>%
    filter(state == state_name) %>%
    ggplot() +
    geom_sf(
      aes(fill = target_homeless_rate_per_10k),
      color = "white",
      linewidth = 0.25
    ) +
    facet_wrap(~ target_year, nrow = 1) +
    scale_fill_viridis_c(
      option = "C",
      limits = rate_limits,
      name = "Homelessness rate\nper 10,000"
    ) +
    coord_sf(datum = NA, expand = FALSE) +
    labs(
      title = paste("Homelessness rates across", state_name, "CoCs"),
      subtitle = "Side-by-side comparison of 2012 and 2025"
    ) +
    theme_void(base_size = 16) +
    theme(
      plot.title = element_text(size = 24, face = "bold"),
      plot.subtitle = element_text(size = 17),
      strip.text = element_text(size = 20, face = "bold"),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 14),
      legend.position = "bottom",
      panel.spacing = grid::unit(1.2, "cm"),
      plot.margin = margin(10, 10, 10, 10)
    )
}

california_map <- make_state_map("California")
florida_map <- make_state_map("Florida")

california_map
florida_map
}
```


```{r}
library(car)        # for Anova()
library(tidyverse)   # dplyr, ggplot2

# Step 1: Read in the data ---------------------------------------------------
dat <- read_excel(
  "outputs/lasso_model/CA_FL_LASSO_MODEL_INPUT_v2.xlsx",
  sheet = "LASSO Model Data"
)

# Step 2: Drop redundant/duplicate ID columns ---------------------------------
# These caused the 76 "singularity" NAs in the earlier fit3 -- each one is just
# a duplicate label for information already captured elsewhere:
#   coc_name         -> same CoC as coc_number, just spelled out
#   state_abbr        -> same state as `state`, just abbreviated
#   target_year       -> always predictor_year + 1, so it's redundant with predictor_year
#   control_time_index / control_state_florida -> re-encodings of predictor_year / state
dat_model <- dat %>%
  select( -coc_number,
    -coc_name,
    -state_abbr,
    -target_year,
    -control_time_index,
    -control_state_florida)

# Step 3: Log-transform the right-skewed predictors ---------------------------
# From the EDA: these coc_-level variables had skewness > 1 (piled low, long
# right tail). Use log(x) when every value is positive and log1p(x) when a
# variable contains genuine zeros. This matches the transformations used by
# the LASSO model below.
data_log <- dat_model

log_vars <- c(
  "coc_population_density_per_sq_mile",
  "coc_real_gdp_per_capita_2017_usd",
  "coc_hic_psh_beds_per_10k",
  "coc_hic_temporary_beds_per_10k",
  "coc_permits_per_1000_housing_units",
  "coc_permits_value_per_1000_housing_units_2025_usd",
  "coc_real_median_household_income_2025_usd",
  "coc_real_per_capita_personal_income_2025_usd",
  "coc_relative_home_price_index_2000_base"
)

# Use log(x) when every value is positive.
# Use log(1 + x) when the variable contains zeros.
for (v in log_vars) {
  if (min(data_log[[v]]) > 0) {
    data_log[[v]] <- log(data_log[[v]])
  } else {
    data_log[[v]] <- log1p(data_log[[v]])
  }
}

# The target is always positive, so use ordinary log()
data_log$target_homeless_rate_per_10k <-
  log(data_log$target_homeless_rate_per_10k)

# Fit the multiple regression using the transformed data
fit_log <- lm(
  target_homeless_rate_per_10k ~ .,
  data = data_log
)

coef_matrix <- summary(fit_log)$coefficients

coef_table <- data.frame(
  Predictor = rownames(coef_matrix),
  Estimate = coef_matrix[, 1],
  Std_Error = coef_matrix[, 2],
  t_value = coef_matrix[, 3],
  p_value = coef_matrix[, 4],
  row.names = NULL
)

coef_table
summary(fit_log)$r.squared
summary(fit_log)$adj.r.squared
summary(fit_log)$sigma
summary(fit_log)$fstatistic

# ---------------------------------------------------------------------------
# Reading the summary() output:
#   Estimate    -- change in log(Y) per 1-unit increase in X (or per log-unit
#                  increase in X, for the predictors we also logged above)
#   Std. Error  -- uncertainty around that estimate
#   t value / Pr(>|t|) -- test of whether that coefficient is significantly
#                          different from 0 (p < .05 = "significant")
#   R-squared   -- proportion of variance in log(Y) explained by the model
#   F-statistic -- tests whether the model as a WHOLE explains more than
#                  just guessing the mean every time
# ---------------------------------------------------------------------------

# Step 6: Confidence intervals for the coefficients -----------------------------
confint(fit_log)   # 95% CI for each beta -- if it crosses 0, not significant

# Step 7: F-test for each individual predictor, holding others fixed -----------
Anova(fit_log)

# Step 8: Model diagnostics on the log scale -------------------------------------
plot(fit_log, 1)   # residuals vs. fitted -- want a random scatter, no funnel shape
plot(fit_log, 2)   # base R's built-in Q-Q version
```
# LASSO

## Why LASSO and not ordinary regression

```{r lasso_setup}
library(glmnet)
library(coefplot)

# Define every object used below so the LASSO section runs in a clean session.
data_all <- dat
target <- "target_homeless_rate_per_10k"
expected_input_md5 <- "5d3fd16b32c687e5207ea59c902e7bef"
actual_input_md5 <- unname(
  tools::md5sum("outputs/lasso_model/CA_FL_LASSO_MODEL_INPUT_v2.xlsx")
)

if (!identical(actual_input_md5, expected_input_md5)) {
  stop(
    "LASSO input MD5 mismatch. Expected ", expected_input_md5,
    "; found ", actual_input_md5
  )
}

if (nrow(data_all) != 887L ||
    length(unique(data_all$coc_number)) != 70L) {
  stop("Unexpected LASSO input shape; expected 887 rows across 70 CoCs.")
}

state_control <- "control_state_florida"
year_control <- "predictor_year"
# This sensitivity run uses year fixed effects rather than the linear time
# index. The original `control_time_index` is deliberately excluded.
controls <- state_control
unpenalized_terms <- c(state_control, year_control)

excluded_columns <- c(
  "state", "state_abbr", "coc_number", "coc_name",
  "predictor_year", "target_year", target, state_control,
  "control_time_index"
)

predictors <- setdiff(names(data_all), excluded_columns)

if (length(predictors) != 38L) {
  stop("Expected 38 candidate predictors; found ", length(predictors), ".")
}

length(predictors)   # should return 38
```

We have 38 predictors, and the EDA found **18 pairs correlated at |r| >= 0.80**.
With that much overlap, ordinary least squares gives unstable coefficients --
small changes in the data swing the estimates around.

LASSO adds a penalty that shrinks coefficients toward zero and sets the least
useful ones **exactly** to zero. The result is a smaller, steadier model.

## Log transformations

We log the target because of the right skew seen above. We also log nine
predictors that are dollar amounts, counts, or index values with long tails.

```{r}
# the nine predictors we are logging
log_vars <- c("coc_population_density_per_sq_mile",
              "coc_real_gdp_per_capita_2017_usd",
              "coc_hic_psh_beds_per_10k",
              "coc_hic_temporary_beds_per_10k",
              "coc_permits_per_1000_housing_units",
              "coc_permits_value_per_1000_housing_units_2025_usd",
              "coc_real_median_household_income_2025_usd",
              "coc_real_per_capita_personal_income_2025_usd",
              "coc_relative_home_price_index_2000_base")

# check the smallest value of each -- log() needs values above zero
sapply(data_all[, log_vars], min)
```

Two of them contain **real zeros**: 60 CoC-years have no permanent-supportive-housing
beds and one has no temporary beds. Those are genuine observations, not missing
data, but `log(0)` is undefined.

The fix is `log1p(x)`, which is just `log(1 + x)`. It turns 0 into 0 and leaves
every other value's ordering unchanged -- no row dropped, no made-up number added.

```{r}
# make a copy so the original data is untouched
data_log <- data_all

for (v in log_vars) {
  if (min(data_log[[v]]) > 0) {
    data_log[[v]] <- log(data_log[[v]])      # safe to log directly
  } else {
    data_log[[v]] <- log1p(data_log[[v]])    # has zeros -> use log(1 + x)
  }
}

# log the target too (its smallest value is 4.83, so plain log is fine)
data_log[[target]] <- log(data_log[[target]])
```

## Building the design matrix

```{r}
# Keep the state control, then convert predictor year into categorical fixed
# effects. 2011 is the reference year; model.matrix creates one indicator for
# each later observed year.
fit_data <- data_log[, c(target, state_control, year_control, predictors)]
fit_data[[year_control]] <- factor(
  fit_data[[year_control]],
  levels = sort(unique(fit_data[[year_control]]))
)

numeric_fit_cols <- vapply(fit_data, is.numeric, logical(1))
if (anyNA(fit_data) ||
    any(!is.finite(as.matrix(fit_data[, numeric_fit_cols, drop = FALSE])))) {
  stop("The categorical-year LASSO inputs contain NA, NaN, or Inf values.")
}

# Y is the response, X is the matrix of predictors.
# model.matrix() builds X and handles any categorical variables for us.
# The [, -1] drops the leading column of 1s, which glmnet adds itself.
Y <- fit_data[[target]]
X <- model.matrix(target_homeless_rate_per_10k ~ ., data = fit_data)[, -1]

year_dummy_cols <- grep(
  paste0("^", year_control),
  colnames(X),
  value = TRUE
)

dim(X)     # 887 rows, 51 columns (38 predictors + state + 12 year indicators)
```

## Choosing lambda by cross validation

`lambda` controls how hard the penalty shrinks. Cross validation estimates the
prediction error at each lambda so we can pick one.

We **force the state control and every categorical year indicator into the
model** by giving them a penalty factor of 0. This adjusts within the observed
years without forcing the relationship with time to be linear.

```{r}
# penalty factor: 0 = never shrink a control, 1 = shrink a candidate predictor
pf <- ifelse(
  colnames(X) %in% c(state_control, year_dummy_cols),
  0,
  1
)

set.seed(10)   # cross validation splits randomly, so we fix the seed
fit_cv <- cv.glmnet(X, Y, alpha = 1, nfolds = 10,
                    intercept = TRUE, penalty.factor = pf)
# alpha = 1 means LASSO (alpha = 0 would be Ridge)

fit_cv$lambda.min    # lambda with the lowest CV error
fit_cv$lambda.1se    # simplest lambda within one standard error of the best
```

```{r lasso_cv_plot, echo=FALSE, out.width="90%"}
# Write the base-R glmnet plot to a file, then insert that file. This avoids
# blank inline graphics devices in Positron.
plot_file <- "lasso_cv_plot.png"

png(
  filename = plot_file,
  width = 1200,
  height = 900,
  res = 150,
  bg = "white"
)

plot(fit_cv)
dev.off()

knitr::include_graphics(plot_file)
```

Reading this plot: red dots are the mean cross-validation error at each lambda,
and the bars are one standard error either side. On the displayed
$-\log(\lambda)$ axis, the left dashed line is `lambda.1se` and the right is
`lambda.min`. The numbers along the top count how
many coefficients are still non-zero.

```{r}
# how many variables does each choice keep?
coef_min <- extract.coef(fit_cv, lambda = "lambda.min")
coef_1se <- extract.coef(fit_cv, lambda = "lambda.1se")

nrow(coef_min) - 1   # includes the unpenalized state and year controls
nrow(coef_1se) - 1
coef_min
```


```{r}
#| fig-height: 4.5
# the coefficient path: each line is one predictor shrinking toward zero
plot(glmnet(X, Y, alpha = 1, penalty.factor = pf), xvar = "lambda", label = TRUE)
```

## Relaxed LASSO

LASSO picks variables but does not give us p-values for its own choices. The
standard fix is to take the variables it picked and refit them with plain `lm()`.
This is called the **relaxed LASSO**.

```{r}
# Keep all prespecified controls, then add only the substantive predictors
# selected by LASSO. The factor is retained as one term, rather than trying to
# use model.matrix's individual dummy-column names in lm().
selected_predictors_min <- intersect(
  setdiff(coef_min$Coefficient, "(Intercept)"),
  predictors
)

# refit those variables with ordinary regression
fit_lm <- lm(
  as.formula(
    paste(
      target,
      "~",
      paste(c(unpenalized_terms, selected_predictors_min), collapse = " + ")
    )
  ),
  data = fit_data
)
summary(fit_lm)
summary(fit_lm)$adj.r.squared
```

```{r}
# Descriptive only: the relaxed refit reuses the data that selected the terms.
sum(summary(fit_lm)$coefficients[-1, 4] < 0.05)
```

These p-values are descriptive only, so backward selection is shown as an
interpretability exercise rather than confirmatory inference.

## Fine-tuning with backward selection

Backward selection: find the least significant substantive predictor, drop it,
refit, and repeat until every removable predictor left is significant. State
and year remain in the model because they were prespecified controls.

```{r}
fit_back <- fit_lm

repeat {
  # Anova() gives one p-value per predictor
  anova_tbl <- Anova(fit_back, type = "II")
  pvals <- anova_tbl[["Pr(>F)"]]
  names(pvals) <- rownames(anova_tbl)

  # Controls stay for adjustment even when their own p-values exceed 0.05.
  removable_pvals <- pvals[
    !is.na(pvals) &
      !names(pvals) %in% c("Residuals", unpenalized_terms)
  ]

  # Stop once every remaining non-control predictor is significant.
  if (length(removable_pvals) == 0 ||
      max(removable_pvals) <= 0.05) break

  worst <- names(removable_pvals)[which.max(removable_pvals)]
  cat("dropping", worst,
      " p =", round(max(removable_pvals), 3), "\n")

  fit_back <- update(fit_back, paste(". ~ . -", worst))
}

fit_final <- fit_back
```

```{r}
length(coef(fit_final)) - 1

summary(fit_final)
summary(fit_final)$adj.r.squared
```

Any terms removed here are a compact descriptive specification, not an
independent validation result. The categorical year effect remains in the model
throughout as a prespecified control.

Worth noticing: **`control_state_florida` remains in the model as a prespecified
control but is not individually significant.** After accounting for the
selected CoC-level variables, the model does not detect a remaining adjusted
state difference.

## Checking the model assumptions

```{r}
#| fig-height: 4.5
plot(fit_final, 1)     # residuals vs fitted -- we want a flat line, even scatter
```

```{r}
#| fig-height: 4.5
plot(fit_final, 2)     # Q-Q plot -- we want points close to the diagonal
```

The residual band is roughly constant across the range, which is exactly what
the log transform was for. On the original rate scale the spread would grow with
the prediction. The Q-Q plot tracks the diagonal through the middle with mild
tail departures -- not perfect, but reasonable.

## Reading the final model

```{r}
# sort the coefficients by how statistically strong they are
final_tbl <- summary(fit_final)$coefficients
final_tbl <- final_tbl[order(-abs(final_tbl[, 3])), ]
round(head(final_tbl, 10), 4)
```

Because the response is logged, a coefficient is roughly the **percent change**
in next-year homelessness for a one-unit rise in the predictor. For the six
predictors that are also logged, both sides are on a log scale, so the
coefficient is an **elasticity** -- a 1% rise in the predictor goes with a beta
percent change in the rate.

**Shelter capacity is the strongest term** (temporary beds, elasticity +0.358).
This does **not** mean shelters cause homelessness. Communities build capacity
where homelessness is already high, and sheltered counts are taken in those
beds -- the association runs from need to capacity.

## A limit worth stating

The p-values above are optimistic. LASSO chose the variables using the same 887
rows that `lm()` then tested them on, and backward selection used those rows
again. Treat them as a guide to which terms are strong, not as formal tests.

Also, `cv.glmnet` shuffles rows at **random**, so a model can be trained partly
on 2024 and evaluated on 2015. That is fine for picking lambda, but it means
nothing above tests whether we can predict a *future* year. Different years
from the same CoC can also appear in both training and validation folds, so the
cross-validation error may be optimistic for predicting entirely new CoCs.

***
## Rolling-origin LASSO

Categorical year fixed effects are not usable for this out-of-time exercise:
the future test year has no estimated year effect. Therefore, the categorical
year sensitivity above is limited to random-fold CV within the observed years.
This retained rolling-origin benchmark reconstructs the original linear time
control so it remains a valid future-year comparison; it is not a test of the
categorical-year specification.

For example, when testing 2025:

1. Fit candidate lambdas using years through 2023.
2. Select lambda using prediction error in 2024.
3. Refit the model using all observations through 2024.
4. Predict 2025.

```{r rolling-origin-lasso}
# Build the original, extrapolatable time-control matrix separately from the
# categorical-year sensitivity matrix used above.
rolling_controls <- c("control_state_florida", "control_time_index")
rolling_fit_data <- data_log[, c(target, rolling_controls, predictors)]
X_rolling <- model.matrix(
  target_homeless_rate_per_10k ~ .,
  data = rolling_fit_data
)[, -1, drop = FALSE]
pf_rolling <- ifelse(colnames(X_rolling) %in% rolling_controls, 0, 1)

rolling_years <- c(
  2017, 2018, 2019, 2020,
  2022, 2023, 2024, 2025
)

rolling_results <- vector(
  "list",
  length(rolling_years)
)

rolling_importance <- vector(
  "list",
  length(rolling_years)
)

for (i in seq_along(rolling_years)) {

  test_year <- rolling_years[i]

  # -----------------------------------------------------------
  # Outer split:
  # Train on every year before the test year.
  # Test only on the selected future year.
  # -----------------------------------------------------------

  outer_train <- which(
    data_all$target_year < test_year
  )

  outer_test <- which(
    data_all$target_year == test_year
  )

  # -----------------------------------------------------------
  # Inner split:
  # Use the latest available training year to select lambda.
  # -----------------------------------------------------------

  years_in_training <- data_all$target_year[outer_train]

  tune_year <- max(years_in_training)

  inner_train <- outer_train[
    years_in_training < tune_year
  ]

  inner_valid <- outer_train[
    years_in_training == tune_year
  ]

  # -----------------------------------------------------------
  # Fit the complete LASSO path using only observations before
  # the lambda-tuning year.
  # -----------------------------------------------------------

  tuning_fit <- glmnet(
    x = X_rolling[inner_train, , drop = FALSE],
    y = Y[inner_train],
    alpha = 1,
    penalty.factor = pf_rolling
  )

  # -----------------------------------------------------------
  # Predict the next chronological year using every candidate
  # lambda produced by glmnet.
  # -----------------------------------------------------------

  tuning_predictions <- predict(
    tuning_fit,
    newx = X_rolling[inner_valid, , drop = FALSE],
    s = tuning_fit$lambda
  )

  # Calculate validation MSE for every candidate lambda.

  tuning_mse <- colMeans(
    sweep(
      tuning_predictions,
      MARGIN = 1,
      STATS = Y[inner_valid],
      FUN = "-"
    )^2
  )

  # Select the lambda with the lowest next-year validation MSE.

  best_lambda <- tuning_fit$lambda[
    which.min(tuning_mse)
  ]

  # -----------------------------------------------------------
  # Refit using every observation before the final test year.
  # -----------------------------------------------------------

  final_rolling_fit <- glmnet(
    x = X_rolling[outer_train, , drop = FALSE],
    y = Y[outer_train],
    alpha = 1,
    lambda = tuning_fit$lambda,
    penalty.factor = pf_rolling
  )

  # -----------------------------------------------------------
  # Predict the completely unseen test year.
  #
  # The LASSO response is log(rate), so exp() returns predictions
  # to homelessness rates per 10,000.
  # -----------------------------------------------------------

  predicted_log_rate <- predict(
    final_rolling_fit,
    newx = X_rolling[outer_test, , drop = FALSE],
    s = best_lambda
  )

  predicted_rate <- exp(
    as.numeric(predicted_log_rate)
  )

  # Count the nonzero coefficients selected in this fold.

  number_nonzero <- sum(
    as.vector(
      coef(final_rolling_fit, s = best_lambda)
    )[-1] != 0
  )

  # Save comparable coefficient importance for this fold.
  fold_coefficients <- as.vector(
    coef(final_rolling_fit, s = best_lambda)
  )[-1]

  training_sd <- apply(
    X_rolling[outer_train, , drop = FALSE],
    2,
    sd
  )

  rolling_importance[[i]] <- data.frame(
    test_year = test_year,
    predictor = colnames(X_rolling),
    coefficient = fold_coefficients,
    standardized_effect = fold_coefficients * training_sd,
    selected = abs(fold_coefficients) > 1e-8
  )

  # Save the results from this test year.

  rolling_results[[i]] <- data.frame(
    test_year = test_year,
    tune_year = tune_year,
    actual = data_all[[target]][outer_test],
    predicted = predicted_rate,
    lambda = best_lambda,
    nonzero_coefficients = number_nonzero
  )
}

# Combine the results from all eight test years.

rolling_lasso <- dplyr::bind_rows(
  rolling_results
)

rolling_importance_all <- dplyr::bind_rows(
  rolling_importance
)

head(rolling_lasso)
```

## Overall rolling-origin results

```{r rolling-lasso-results}
rolling_lasso_score <- data.frame(

  observations = nrow(rolling_lasso),

  RMSE = sqrt(
    mean(
      (rolling_lasso$actual -
         rolling_lasso$predicted)^2
    )
  ),

  MAE = mean(
    abs(
      rolling_lasso$actual -
        rolling_lasso$predicted
    )
  ),

  R2 = 1 -
    sum(
      (rolling_lasso$actual -
         rolling_lasso$predicted)^2
    ) /
    sum(
      (rolling_lasso$actual -
         mean(rolling_lasso$actual))^2
    )
)

round(rolling_lasso_score, 3)
```

The rolling-origin LASSO is evaluated on 555 future-year observations. Its
overall results are approximately:

- RMSE: 16.50
- MAE: 10.07
- Predictive R-squared: 0.542

These are out-of-time results: neither coefficient fitting nor lambda selection
uses observations from the final test year.

## Rolling-origin results by year

```{r rolling-lasso-by-year}
rolling_lasso_by_year <- rolling_lasso %>%
  group_by(test_year, tune_year) %>%
  summarise(
    observations = n(),

    RMSE = sqrt(
      mean((actual - predicted)^2)
    ),

    lambda = first(lambda),

    nonzero_coefficients =
      first(nonzero_coefficients),

    .groups = "drop"
  )

rolling_lasso_by_year
```

## Factors selected by the rolling LASSO

Because rolling-origin evaluation fits eight models, there is no single
coefficient table. The most stable predictors are the ones selected in the
largest number of test-year folds. Standardized effects make coefficient
magnitudes comparable across predictors measured in different units.

```{r rolling-lasso-importance}
importance_table <- rolling_importance_all %>%
  filter(!predictor %in% controls) %>%
  group_by(predictor) %>%
  summarise(
    folds_selected = sum(selected),
    selection_rate = mean(selected),
    mean_abs_standardized_effect = if (any(selected)) {
      mean(abs(standardized_effect[selected]))
    } else {
      0
    },
    average_direction = case_when(
      mean(standardized_effect) > 0 ~ "Positive",
      mean(standardized_effect) < 0 ~ "Negative",
      TRUE ~ "None"
    ),
    .groups = "drop"
  ) %>%
  arrange(
    desc(folds_selected),
    desc(mean_abs_standardized_effect)
  )

importance_table
```

`folds_selected` ranges from 0 to 8. Predictors selected in most or all folds
are the most consistently useful for future-year prediction. Direction
describes association, not causation.

# Neural Network

## What changes for a neural network

A neural network can capture curved and interacting relationships that a linear
model cannot. Four things differ from the textbook classification example:

| Textbook setup | What we do here | Why |
|---|---|---|
| Classification, softmax output | **Regression**: one linear output unit, MSE loss | Our target is a continuous rate |
| Random train/test split | **Rolling origin** over 8 future years | Random splits leak future information |
| Random internal validation split | Hold out the most recent training year | Same reason |
| No scaling | Scale using **training rows only** | The model must not see the validation year |

```{r}
#| eval: false
# Keras runs on Python. This loads the environment set up in Module 7.
library(reticulate)
use_condaenv("r", required = TRUE)
library(keras3)
```

```{r}
#| include: false
suppressMessages({
  library(reticulate)
  try(use_condaenv("r", required = TRUE), silent = TRUE)
  library(keras3)
})
```

## Setting up rolling-origin validation

Instead of one random split, we validate **eight times**. Each time the model
trains only on years before the validation year, then predicts that year once.

```{r}
# the eight years we validate on (2021 is absent -- COVID)
valid_years <- rolling_years

# The neural-net benchmark retains its two numerical controls; it is separate
# from the categorical-year LASSO sensitivity above.
nn_controls <- c("control_state_florida", "control_time_index")
nn_features <- c(nn_controls, predictors)

# use the ORIGINAL (unlogged) data here, so results compare to the LASSO
X_all <- as.matrix(data_all[, nn_features])
y_all <- data_all[[target]]

# show how the training window grows
for (vy in valid_years) {
  cat("validate", vy, "-> train on", sum(data_all$target_year < vy),
      "rows, score", sum(data_all$target_year == vy), "rows\n")
}
```

## Building a network

```{r}
# A small helper so we can try different shapes without repeating ourselves.
# units = how many neurons in each hidden layer.
build_net <- function(n_inputs, units, dropout = 0) {
  model <- keras_model_sequential(input_shape = c(n_inputs))

  for (u in units) {
    model <- model %>% layer_dense(units = u, activation = "relu")
    if (dropout > 0) model <- model %>% layer_dropout(rate = dropout)
  }

  # ONE output unit with no activation -- this makes it a regression
  model <- model %>% layer_dense(units = 1)

  model %>% compile(optimizer = "rmsprop", loss = "mse", metrics = c("mae"))
  model
}
```

## Fitting one fold

This function does the whole job for a single validation year: scale on the
training rows, decide how many epochs to use, refit, and predict.

```{r}
fit_one_fold <- function(valid_year, units, dropout, seed) {

  train_rows <- which(data_all$target_year <  valid_year)
  valid_rows <- which(data_all$target_year == valid_year)

  # --- scaling, computed on TRAINING rows only ---
  mu <- colMeans(X_all[train_rows, ])
  sdv <- apply(X_all[train_rows, ], 2, sd)
  sdv[sdv == 0] <- 1                       # avoid dividing by zero

  x_train <- scale(X_all[train_rows, ], center = mu, scale = sdv)
  x_valid <- scale(X_all[valid_rows, ], center = mu, scale = sdv)

  # scale the target too, then undo it on the predictions later
  y_mu <- mean(y_all[train_rows]); y_sd <- sd(y_all[train_rows])
  y_train <- (y_all[train_rows] - y_mu) / y_sd

  # --- inner split: hold out the most recent training year ---
  # (a random split here would let future years influence when we stop)
  yrs <- data_all$target_year[train_rows]
  inner_train <- which(yrs <  max(yrs))
  inner_valid <- which(yrs == max(yrs))

  set_random_seed(seed)

  # fit while watching the held-out year, and stop when it stops improving
  m0 <- build_net(ncol(x_train), units, dropout)
  history <- m0 %>% fit(
    x_train[inner_train, ], y_train[inner_train],
    validation_data = list(x_train[inner_valid, ], y_train[inner_valid]),
    epochs = 300, batch_size = 32, verbose = 0,
    callbacks = list(callback_early_stopping(monitor = "val_loss",
                                             patience = 20,
                                             restore_best_weights = TRUE)))

  best_epochs <- which.min(history$metrics$val_loss)

  # --- refit on the full training window for that many epochs ---
  set_random_seed(seed)
  m <- build_net(ncol(x_train), units, dropout)
  m %>% fit(x_train, y_train, epochs = best_epochs,
            batch_size = 32, verbose = 0)

  # predict, then undo the target scaling
  pred <- as.numeric(predict(m, x_valid, verbose = 0)) * y_sd + y_mu

  data.frame(valid_year = valid_year,
             actual = y_all[valid_rows],
             predicted = pred)
}
```

## Running all eight folds

Network training is random, so on a panel this small a single run is mostly
noise. We fit each fold under **five different seeds** and average the
predictions.

```{r}
#| eval: false
# This takes roughly 8 minutes for one architecture (8 folds x 5 seeds,
# about 11 seconds each). Set eval: true to actually run it.
seeds <- c(10, 42, 123, 2024, 31337)

run_network <- function(units, dropout) {
  out <- list()
  for (vy in valid_years) {
    # fit the fold once per seed
    folds <- lapply(seeds, function(s) fit_one_fold(vy, units, dropout, s))

    # the actual values are identical every time, so keep the first result
    # and replace its predictions with the average across all five seeds
    result <- folds[[1]]
    result$predicted <- rowMeans(sapply(folds, function(f) f$predicted))

    out[[length(out) + 1]] <- result
  }
  do.call(rbind, out)
}

nn_16_8   <- run_network(units = c(16, 8), dropout = 0)     # standard shape
nn_linear <- run_network(units = c(),      dropout = 0)     # no hidden layers
```

## Scoring the results

We use the same three measures throughout the project, so every model can be
compared on equal terms.

```{r}
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae  <- function(actual, pred) mean(abs(actual - pred))
r2   <- function(actual, pred) 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)

score <- function(df, label) {
  data.frame(model = label,
             n     = nrow(df),
             RMSE  = rmse(df$actual, df$predicted),
             MAE   = mae(df$actual, df$predicted),
             R2    = r2(df$actual, df$predicted))
}
```

```{r}
#| eval: false
rbind(score(nn_16_8,   "neural net 16-8"),
      score(nn_linear, "no hidden layers (= OLS)"))
```

Running the full set of five architectures (in `fit_neural_net.R`) gives:

| Model | RMSE | MAE | R² |
|---|---:|---:|---:|
| Neural net, 32-16 with dropout | **12.64** | 8.43 | **0.731** |
| Neural net, 16-8 | 12.92 | 9.11 | 0.719 |
| Neural net, no hidden layers (= OLS) | 16.06 | 11.46 | 0.565 |
| Rolling-origin LASSO | 16.50 | 10.07 | 0.542 |
| State + time baseline | 21.93 | 14.98 | 0.190 |

Two things to read off this table.

**The hidden-layer networks beat the rolling LASSO.** The best network reduces
RMSE by about 23% relative to the time-ordered LASSO.

**The no-hidden-layer version helps identify the benefit of hidden layers.**
It has RMSE 16.06, compared with 12.64 for the best hidden-layer network under
the same neural-network evaluation setup. That improvement supports the value
of nonlinear relationships. The linear network and LASSO are close, and they
should not be treated as numerically identical because they use different
penalties and training procedures.

## The catch

```{r}
#| eval: false
# look at how much the answer moves between seeds for a single fold
sapply(seeds, function(s) {
  f <- fit_one_fold(2025, units = c(16, 8), dropout = 0, seed = s)
  rmse(f$actual, f$predicted)
})
```

Across the whole run, a single 16-8 network scores anywhere from **14.38 to
16.29** depending only on the random seed. The five-seed average scores 12.92 --
better than *any* individual seed. So a large part of the improvement is
averaging away instability, not one network being better.

The weakest seed only narrowly beats the rolling LASSO (16.29 versus 16.50).
The LASSO, by contrast, is deterministic here: same data, same answer, every
time.

The honest statement is: **the ensemble clearly beats the LASSO; individual
networks are usually better here but substantially less stable.**

## The result that matters most

We also asked whether either model finds anything beyond simply knowing **last
year's rate** for that CoC.

| Model | RMSE | R² |
|---|---:|---:|
| LASSO: state + time + prior rate | **8.89** | 0.866 |
| Neural net on prior rate alone | 8.93 | 0.865 |
| Prior rate only | 9.12 | 0.859 |
| Neural net: prior rate + 38 predictors | 10.05 | 0.829 |
| LASSO: prior rate + 38 predictors | 8.89 | 0.866 |

Given the prior year's rate, **the LASSO shrank all 38 predictors to exactly
zero** -- in every fold, under both lambda rules. And **adding those 38
predictors made the neural network worse**, from 8.93 up to 10.05.

The reason is the difference between the two methods. LASSO can say "ignore
these" and pay nothing for having been offered them. A neural network cannot --
every input stays wired into the first layer, so 38 uninformative predictors
inject noise it has no way to discard.

Two model families with completely different failure modes reached the same
conclusion: **next-year homelessness is dominated by its own prior level, and
the predictor set adds little on top of that.**

***

# Summary {-}

```{r}
#| echo: false
data.frame(
  Stage = c("Data", "Full-sample LASSO", "Rolling LASSO",
            "Neural net", "Both"),
  Result = c("887 CoC-years, 70 CoCs, 38 predictors, 0 missing values",
             "Categorical year sensitivity: 23 candidates at lambda.min, 11 at lambda.1se",
             "Out-of-time RMSE 16.50, R-squared 0.542",
             "Best out-of-time RMSE 12.64, R-squared 0.731",
             "Predictors add nothing once last year's rate is known"))
```

Everything above describes **association, not causation**. Two states observed
over a short, strongly time-ordered panel cannot support causal claims, and the
shelter-bed result is the clearest reminder of why: it is one of the strongest
predictors in every model, and it almost certainly reflects capacity following
need rather than the other way round.

# Reproduce {-}

```{r}
#| eval: false
# the full production versions of these three stages
source("scripts/build_expanded_lasso_input_v2.R")   # data
source("scripts/fit_lasso_logged.R")                # LASSO
source("scripts/fit_neural_net.R")                  # neural network (~35 min)
```
