---
title: "CA/FL Homelessness: Data, LASSO, and Neural Network"
author: "Data Science Group Project"
format:
  html:
    code-fold: show
    highlight-style: haddock
    number-sections: true
    theme: lumen
    toc: true
    toc-depth: 4
    toc-location: left
  pdf:
    number-sections: true
    toc: true
    toc-depth: 4
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
urlcolor: blue
editor_options:
  chunk_output_type: inline
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 4)
options(scipen = 0, digits = 4)   # controls base R output

# the project keeps its packages in a folder called _r_libs so that everyone
# on the team is running the same versions
.libPaths(c(file.path(getwd(), "_r_libs"), .libPaths()))

if (!require("pacman")) install.packages("pacman")
pacman::p_load(openxlsx, dplyr, ggplot2, glmnet, coefplot, car)
```

# Objectives {-}

This file reproduces three stages of the project:

1. **Data acquisition and cleaning** -- where the modeling table comes from and what
   was checked before any model was fitted.
2. **LASSO** -- variable selection on a log-transformed target, followed by a
   relaxed refit and backward selection.
3. **Neural network** -- a feed-forward benchmark validated on future years the
   model has never seen.

Every number produced here matches the project report. All results are
**associations, not causes**.

---

# Data Acquisition and Cleaning

## Where the data comes from

The outcome and the predictors are published on two different maps, and
reconciling them is the main data task in this project.

| What | Source | Published at |
|---|---|---|
| Homelessness counts (the outcome) | HUD Point-in-Time (PIT) counts | **Continuum of Care** |
| Shelter / housing capacity | HUD Housing Inventory Count (HIC) | Continuum of Care |
| Population, poverty, income, housing | U.S. Census / ACS, BEA, BLS | **County** |
| Home prices | FHFA, Zillow | County |
| Policy (min wage, Medicaid, TANF, SSI) | State sources, KFF, CBPP | State |

A **Continuum of Care (CoC)** is HUD's regional planning unit for homelessness.
It might be one big county, several rural counties together, or a city carved
out of the county around it. HUD counts homelessness there, so that is the level
we have to model at.

County predictors were therefore allocated onto CoC geography using HUD's
FY2024 county-to-CoC crosswalk. Where a county is split between two CoCs (this
happens for 8 CoCs), its values are divided using census-tract population
shares.

That whole build is long, so it lives in its own scripts rather than here:

```{r}
#| eval: false
# Not run here -- these are the scripts that build the modeling table.
# Run them only if you need to rebuild the data from the raw downloads.
source("scripts/build_county_raw_panel.R")      # ~2,000 county-years from raw sources
source("scripts/build_coc_lasso_panel.R")       # county -> CoC allocation
source("scripts/build_expanded_lasso_input_v2.R")  # final audited modeling table
```

## Loading the modeling table

```{r}
# The final modeling table. One row = one CoC in one year.
data_all <- read.xlsx("outputs/lasso_model/CA_FL_LASSO_MODEL_INPUT_v2.xlsx",
                      sheet = "LASSO Model Data")

dim(data_all)          # expect 887 rows and 47 columns
```

The 47 columns break into four groups. Keeping them straight matters, because
the identifier columns must **never** be used as predictors.

```{r}
# 6 identifier columns -- labels only, never predictors
id_cols <- c("state", "state_abbr", "coc_number", "coc_name",
             "predictor_year", "target_year")

# the thing we are predicting
target <- "target_homeless_rate_per_10k"

# 2 controls: always kept in the model so we adjust for state and year
controls <- c("control_state_florida", "control_time_index")

# everything else is a candidate predictor
predictors <- setdiff(names(data_all), c(id_cols, target, controls))

length(predictors)     # expect 38
```

## Understanding one row

```{r}
# look at a single CoC to see how the rows are laid out
data_all %>%
  filter(coc_number == "CA-501") %>%
  select(coc_name, predictor_year, target_year, target_homeless_rate_per_10k) %>%
  head(4)
```

Notice `target_year` is always `predictor_year + 1`. **The predictors describe
year t and the outcome is year t+1.** This is deliberate: the model is always
predicting something that had not happened yet when the predictors were
measured, which rules out the most obvious way a model can accidentally cheat.

## Cleaning checks

The cleaning rule for this project is simple: **a missing value stays missing.**
Nothing is filled with a zero, an average, or an interpolated guess. So before
modeling we confirm there is nothing missing left.

```{r}
# count missing values across the target, controls, and predictors
model_cols <- c(target, controls, predictors)
sum(is.na(data_all[, model_cols]))       # expect 0

# also check for infinite values, which NA checks miss
sum(!is.finite(as.matrix(data_all[, model_cols])))   # expect 0
```

```{r}
# how many CoCs, and which years do we have?
length(unique(data_all$coc_number))      # expect 70
sort(unique(data_all$target_year))       # 2012-2025, with 2021 missing
```

**2021 is absent on purpose.** COVID disrupted the Point-in-Time count so badly
that the number is not comparable with the years around it, so it is excluded as
a modeling target. That is why we have 13 usable years, not 14.

```{r}
# confirm every row is a unique CoC-year (no accidental duplicates)
sum(duplicated(data_all[, c("coc_number", "target_year")]))   # expect 0

# confirm the time ordering holds for every row
all(data_all$target_year == data_all$predictor_year + 1)      # expect TRUE
```

## What the outcome looks like

```{r}
#| fig-height: 3.5
summary(data_all$target_homeless_rate_per_10k)

ggplot(data_all, aes(x = target_homeless_rate_per_10k)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(title = "Next-year homeless rate per 10k",
       x = "rate per 10,000 residents", y = "count") +
  theme_bw()
```

The distribution has a long right tail. Most CoC-years sit between roughly 10
and 25 per 10k, but a few stretch past 100. This right skew is what motivates
the log transform in the next section.

```{r}
# compare the two states
data_all %>%
  group_by(state) %>%
  summarise(n = n(),
            mean_rate = mean(target_homeless_rate_per_10k),
            sd_rate   = sd(target_homeless_rate_per_10k))
```

California averages about **twice** Florida's rate across the full panel
(roughly 41 vs 21 per 10k) -- this gap is the whole motivation for the study.

A note on comparing to the report: the report quotes 42.6 vs 16.2 per 10k. Those
are computed on the **555 validation rows only** (the eight future years the
models are scored on), not all 887 rows. Both are correct -- they just cover
different sets of years, and Florida's rate falls over time, which is why its
average is lower in the later years.

---

# LASSO

## Why LASSO and not ordinary regression

We have 38 predictors, and the EDA found **18 pairs correlated at |r| >= 0.80**.
With that much overlap, ordinary least squares gives unstable coefficients --
small changes in the data swing the estimates around.

LASSO adds a penalty that shrinks coefficients toward zero and sets the least
useful ones **exactly** to zero. The result is a smaller, steadier model.

## Log transformations

We log the target because of the right skew seen above. We also log nine
predictors that are dollar amounts, counts, or index values with long tails.

```{r}
# the nine predictors we are logging
log_vars <- c("coc_population_density_per_sq_mile",
              "coc_real_gdp_per_capita_2017_usd",
              "coc_hic_psh_beds_per_10k",
              "coc_hic_temporary_beds_per_10k",
              "coc_permits_per_1000_housing_units",
              "coc_permits_value_per_1000_housing_units_2025_usd",
              "coc_real_median_household_income_2025_usd",
              "coc_real_per_capita_personal_income_2025_usd",
              "coc_relative_home_price_index_2000_base")

# check the smallest value of each -- log() needs values above zero
sapply(data_all[, log_vars], min)
```

Two of them contain **real zeros**: 60 CoC-years have no permanent-supportive-housing
beds and one has no temporary beds. Those are genuine observations, not missing
data, but `log(0)` is undefined.

The fix is `log1p(x)`, which is just `log(1 + x)`. It turns 0 into 0 and leaves
every other value's ordering unchanged -- no row dropped, no made-up number added.

```{r}
# make a copy so the original data is untouched
data_log <- data_all

for (v in log_vars) {
  if (min(data_log[[v]]) > 0) {
    data_log[[v]] <- log(data_log[[v]])      # safe to log directly
  } else {
    data_log[[v]] <- log1p(data_log[[v]])    # has zeros -> use log(1 + x)
  }
}

# log the target too (its smallest value is 4.83, so plain log is fine)
data_log[[target]] <- log(data_log[[target]])
```

## Building the design matrix

```{r}
# keep only the columns that go into the model
fit_data <- data_log[, c(target, controls, predictors)]

# Y is the response, X is the matrix of predictors.
# model.matrix() builds X and handles any categorical variables for us.
# The [, -1] drops the leading column of 1s, which glmnet adds itself.
Y <- fit_data[[target]]
X <- model.matrix(target_homeless_rate_per_10k ~ ., data = fit_data)[, -1]

dim(X)     # 887 rows, 40 columns (38 predictors + 2 controls)
```

## Choosing lambda by cross validation

`lambda` controls how hard the penalty shrinks. Cross validation estimates the
prediction error at each lambda so we can pick one.

We also **force the two controls into the model** by giving them a penalty
factor of 0, so state and year are always adjusted for no matter how strong the
penalty gets.

```{r}
# penalty factor: 0 = never shrink this variable, 1 = shrink normally
pf <- ifelse(colnames(X) %in% controls, 0, 1)

set.seed(10)   # cross validation splits randomly, so we fix the seed
fit_cv <- cv.glmnet(X, Y, alpha = 1, nfolds = 10,
                    intercept = TRUE, penalty.factor = pf)
# alpha = 1 means LASSO (alpha = 0 would be Ridge)

fit_cv$lambda.min    # lambda with the lowest CV error
fit_cv$lambda.1se    # simplest lambda within one standard error of the best
```

```{r}
#| fig-height: 4.5
plot(fit_cv)
```

Reading this plot: red dots are the mean cross-validation error at each lambda,
and the bars are one standard error either side. The left dashed line is
`lambda.min`, the right is `lambda.1se`. The numbers along the top count how
many coefficients are still non-zero.

```{r}
# how many variables does each choice keep?
coef_min <- extract.coef(fit_cv, lambda = "lambda.min")
coef_1se <- extract.coef(fit_cv, lambda = "lambda.1se")

nrow(coef_min) - 1   # minus 1 for the intercept -- expect 27
nrow(coef_1se) - 1   # expect 16
```

```{r}
#| fig-height: 4.5
# the coefficient path: each line is one predictor shrinking toward zero
plot(glmnet(X, Y, alpha = 1, penalty.factor = pf), xvar = "lambda", label = TRUE)
```

## Relaxed LASSO

LASSO picks variables but does not give us p-values for its own choices. The
standard fix is to take the variables it picked and refit them with plain `lm()`.
This is called the **relaxed LASSO**.

```{r}
# the variable names LASSO selected at lambda.min
var_min <- setdiff(coef_min$Coefficient, "(Intercept)")

# refit those variables with ordinary regression
fit_lm <- lm(as.formula(paste(target, "~", paste(var_min, collapse = " + "))),
             data = fit_data)

summary(fit_lm)$adj.r.squared      # expect 0.6436
```

```{r}
# how many of the 27 are actually significant?
sum(summary(fit_lm)$coefficients[-1, 4] < 0.05)   # expect 18
```

Only 18 of 27 clear the 0.05 bar, so there is still trimming to do.

## Fine-tuning with backward selection

Backward selection: find the least significant predictor, drop it, refit, and
repeat until everything left is significant.

```{r}
fit_back <- fit_lm

repeat {
  # Anova() gives one p-value per predictor
  pvals <- Anova(fit_back, type = "II")[["Pr(>F)"]]
  names(pvals) <- rownames(Anova(fit_back, type = "II"))
  pvals <- pvals[!is.na(pvals) & names(pvals) != "Residuals"]

  # stop once every remaining predictor is significant
  if (max(pvals) <= 0.05) break

  worst <- names(pvals)[which.max(pvals)]
  cat("dropping", worst, " p =", round(max(pvals), 3), "\n")

  fit_back <- update(fit_back, paste(". ~ . -", worst))
}

fit_final <- fit_back
```

```{r}
length(coef(fit_final)) - 1              # expect 20 predictors
summary(fit_final)$adj.r.squared         # expect 0.6449
```

Seven predictors came out and the adjusted R-squared barely moved (0.6436 to
0.6449). That is the point -- those seven were costing us model complexity
without adding explanation.

Worth noticing: **`control_state_florida` was forced into the LASSO but backward
selection dropped it.** Once the CoC-level variables are in the model, the state
indicator has nothing left to explain.

## Checking the model assumptions

```{r}
#| fig-height: 4.5
plot(fit_final, 1)     # residuals vs fitted -- we want a flat line, even scatter
```

```{r}
#| fig-height: 4.5
plot(fit_final, 2)     # Q-Q plot -- we want points close to the diagonal
```

The residual band is roughly constant across the range, which is exactly what
the log transform was for. On the original rate scale the spread would grow with
the prediction. The Q-Q plot tracks the diagonal through the middle with mild
tail departures -- not perfect, but reasonable.

## Reading the final model

```{r}
# sort the coefficients by how statistically strong they are
final_tbl <- summary(fit_final)$coefficients
final_tbl <- final_tbl[order(-abs(final_tbl[, 3])), ]
round(head(final_tbl, 10), 4)
```

Because the response is logged, a coefficient is roughly the **percent change**
in next-year homelessness for a one-unit rise in the predictor. For the six
predictors that are also logged, both sides are on a log scale, so the
coefficient is an **elasticity** -- a 1% rise in the predictor goes with a beta
percent change in the rate.

**Shelter capacity is the strongest term** (temporary beds, elasticity +0.358).
This does **not** mean shelters cause homelessness. Communities build capacity
where homelessness is already high, and sheltered counts are taken in those
beds -- the association runs from need to capacity.

## A limit worth stating

The p-values above are optimistic. LASSO chose the variables using the same 887
rows that `lm()` then tested them on, and backward selection used those rows
again. Treat them as a guide to which terms are strong, not as formal tests.

Also, `cv.glmnet` shuffles rows at **random**, so a model can be trained partly
on 2024 and evaluated on 2015. That is fine for picking lambda, but it means
nothing above tests whether we can predict a *future* year. The next section
does exactly that.

---

# Neural Network

## What changes for a neural network

A neural network can capture curved and interacting relationships that a linear
model cannot. Four things differ from the textbook classification example:

| Textbook setup | What we do here | Why |
|---|---|---|
| Classification, softmax output | **Regression**: one linear output unit, MSE loss | Our target is a continuous rate |
| Random train/test split | **Rolling origin** over 8 future years | Random splits leak future information |
| Random internal validation split | Hold out the most recent training year | Same reason |
| No scaling | Scale using **training rows only** | The model must not see the validation year |

```{r}
#| eval: false
# Keras runs on Python. This loads the environment set up in Module 7.
library(reticulate)
use_condaenv("r", required = TRUE)
library(keras3)
```

```{r}
#| include: false
suppressMessages({
  library(reticulate)
  try(use_condaenv("r", required = TRUE), silent = TRUE)
  library(keras3)
})
```

## Setting up rolling-origin validation

Instead of one random split, we validate **eight times**. Each time the model
trains only on years before the validation year, then predicts that year once.

```{r}
# the eight years we validate on (2021 is absent -- COVID)
valid_years <- c(2017, 2018, 2019, 2020, 2022, 2023, 2024, 2025)

# the columns the network sees: 2 controls + 38 predictors = 40 inputs
nn_features <- c(controls, predictors)

# use the ORIGINAL (unlogged) data here, so results compare to the LASSO
X_all <- as.matrix(data_all[, nn_features])
y_all <- data_all[[target]]

# show how the training window grows
for (vy in valid_years) {
  cat("validate", vy, "-> train on", sum(data_all$target_year < vy),
      "rows, score", sum(data_all$target_year == vy), "rows\n")
}
```

## Building a network

```{r}
# A small helper so we can try different shapes without repeating ourselves.
# units = how many neurons in each hidden layer.
build_net <- function(n_inputs, units, dropout = 0) {
  model <- keras_model_sequential(input_shape = c(n_inputs))

  for (u in units) {
    model <- model %>% layer_dense(units = u, activation = "relu")
    if (dropout > 0) model <- model %>% layer_dropout(rate = dropout)
  }

  # ONE output unit with no activation -- this makes it a regression
  model <- model %>% layer_dense(units = 1)

  model %>% compile(optimizer = "rmsprop", loss = "mse", metrics = c("mae"))
  model
}
```

## Fitting one fold

This function does the whole job for a single validation year: scale on the
training rows, decide how many epochs to use, refit, and predict.

```{r}
fit_one_fold <- function(valid_year, units, dropout, seed) {

  train_rows <- which(data_all$target_year <  valid_year)
  valid_rows <- which(data_all$target_year == valid_year)

  # --- scaling, computed on TRAINING rows only ---
  mu <- colMeans(X_all[train_rows, ])
  sdv <- apply(X_all[train_rows, ], 2, sd)
  sdv[sdv == 0] <- 1                       # avoid dividing by zero

  x_train <- scale(X_all[train_rows, ], center = mu, scale = sdv)
  x_valid <- scale(X_all[valid_rows, ], center = mu, scale = sdv)

  # scale the target too, then undo it on the predictions later
  y_mu <- mean(y_all[train_rows]); y_sd <- sd(y_all[train_rows])
  y_train <- (y_all[train_rows] - y_mu) / y_sd

  # --- inner split: hold out the most recent training year ---
  # (a random split here would let future years influence when we stop)
  yrs <- data_all$target_year[train_rows]
  inner_train <- which(yrs <  max(yrs))
  inner_valid <- which(yrs == max(yrs))

  set_random_seed(seed)

  # fit while watching the held-out year, and stop when it stops improving
  m0 <- build_net(ncol(x_train), units, dropout)
  history <- m0 %>% fit(
    x_train[inner_train, ], y_train[inner_train],
    validation_data = list(x_train[inner_valid, ], y_train[inner_valid]),
    epochs = 300, batch_size = 32, verbose = 0,
    callbacks = list(callback_early_stopping(monitor = "val_loss",
                                             patience = 20,
                                             restore_best_weights = TRUE)))

  best_epochs <- which.min(history$metrics$val_loss)

  # --- refit on the full training window for that many epochs ---
  set_random_seed(seed)
  m <- build_net(ncol(x_train), units, dropout)
  m %>% fit(x_train, y_train, epochs = best_epochs,
            batch_size = 32, verbose = 0)

  # predict, then undo the target scaling
  pred <- as.numeric(predict(m, x_valid, verbose = 0)) * y_sd + y_mu

  data.frame(valid_year = valid_year,
             actual = y_all[valid_rows],
             predicted = pred)
}
```

## Running all eight folds

Network training is random, so on a panel this small a single run is mostly
noise. We fit each fold under **five different seeds** and average the
predictions.

```{r}
#| eval: false
# This takes roughly 8 minutes for one architecture (8 folds x 5 seeds,
# about 11 seconds each). Set eval: true to actually run it.
seeds <- c(10, 42, 123, 2024, 31337)

run_network <- function(units, dropout) {
  out <- list()
  for (vy in valid_years) {
    # fit the fold once per seed
    folds <- lapply(seeds, function(s) fit_one_fold(vy, units, dropout, s))

    # the actual values are identical every time, so keep the first result
    # and replace its predictions with the average across all five seeds
    result <- folds[[1]]
    result$predicted <- rowMeans(sapply(folds, function(f) f$predicted))

    out[[length(out) + 1]] <- result
  }
  do.call(rbind, out)
}

nn_16_8   <- run_network(units = c(16, 8), dropout = 0)     # standard shape
nn_linear <- run_network(units = c(),      dropout = 0)     # no hidden layers
```

## Scoring the results

We use the same three measures throughout the project, so every model can be
compared on equal terms.

```{r}
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae  <- function(actual, pred) mean(abs(actual - pred))
r2   <- function(actual, pred) 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)

score <- function(df, label) {
  data.frame(model = label,
             n     = nrow(df),
             RMSE  = rmse(df$actual, df$predicted),
             MAE   = mae(df$actual, df$predicted),
             R2    = r2(df$actual, df$predicted))
}
```

```{r}
#| eval: false
rbind(score(nn_16_8,   "neural net 16-8"),
      score(nn_linear, "no hidden layers (= OLS)"))
```

Running the full set of five architectures (in `fit_neural_net.R`) gives:

| Model | RMSE | MAE | R² |
|---|---:|---:|---:|
| Neural net, 32-16 with dropout | **12.64** | 8.43 | **0.731** |
| Neural net, 16-8 | 12.92 | 9.11 | 0.719 |
| LASSO with state interactions | 14.69 | 10.14 | 0.636 |
| Neural net, no hidden layers (= OLS) | 16.06 | 11.46 | 0.565 |
| State + time baseline | 21.93 | 14.98 | 0.190 |

Two things to read off this table.

**The networks beat the LASSO.** The best is a 14% reduction in RMSE.

**The no-hidden-layer version tells us why.** With no hidden layers the network
*is* ordinary least squares, and it does **worse** than the LASSO (16.06 vs
14.69) -- exactly as expected, since that is the situation the LASSO penalty is
built for. So the network's advantage comes from **nonlinearity**, not from
anything else in how it was set up.

## The catch

```{r}
#| eval: false
# look at how much the answer moves between seeds for a single fold
sapply(seeds, function(s) {
  f <- fit_one_fold(2025, units = c(16, 8), dropout = 0, seed = s)
  rmse(f$actual, f$predicted)
})
```

Across the whole run, a single 16-8 network scores anywhere from **14.38 to
16.29** depending only on the random seed. The five-seed average scores 12.92 --
better than *any* individual seed. So a large part of the improvement is
averaging away instability, not one network being better.

An unlucky seed loses to the LASSO outright. The LASSO, by contrast, is
deterministic: same data, same answer, every time.

The honest statement is: **an ensemble of five networks beats the LASSO; a
single network is better on average but unreliable.**

## The result that matters most

We also asked whether either model finds anything beyond simply knowing **last
year's rate** for that CoC.

| Model | RMSE | R² |
|---|---:|---:|
| LASSO: state + time + prior rate | **8.89** | 0.866 |
| Neural net on prior rate alone | 8.93 | 0.865 |
| Prior rate only | 9.12 | 0.859 |
| Neural net: prior rate + 38 predictors | 10.05 | 0.829 |
| LASSO: prior rate + 38 predictors | 8.89 | 0.866 |

Given the prior year's rate, **the LASSO shrank all 38 predictors to exactly
zero** -- in every fold, under both lambda rules. And **adding those 38
predictors made the neural network worse**, from 8.93 up to 10.05.

The reason is the difference between the two methods. LASSO can say "ignore
these" and pay nothing for having been offered them. A neural network cannot --
every input stays wired into the first layer, so 38 uninformative predictors
inject noise it has no way to discard.

Two model families with completely different failure modes reached the same
conclusion: **next-year homelessness is dominated by its own prior level, and
the predictor set adds little on top of that.**

---

# Summary {-}

```{r}
#| echo: false
data.frame(
  Stage = c("Data", "LASSO", "LASSO", "Neural net", "Both"),
  Result = c("887 CoC-years, 70 CoCs, 38 predictors, 0 missing values",
             "27 variables at lambda.min, 16 at lambda.1se",
             "Final model: 20 predictors, adjusted R-squared 0.645",
             "Best out-of-time RMSE 12.64, R-squared 0.731",
             "Predictors add nothing once last year's rate is known"))
```

Everything above describes **association, not causation**. Two states observed
over a short, strongly time-ordered panel cannot support causal claims, and the
shelter-bed result is the clearest reminder of why: it is one of the strongest
predictors in every model, and it almost certainly reflects capacity following
need rather than the other way round.

# Reproduce {-}

```{r}
#| eval: false
# the full production versions of these three stages
source("scripts/build_expanded_lasso_input_v2.R")   # data
source("scripts/fit_lasso_logged.R")                # LASSO
source("scripts/fit_neural_net.R")                  # neural network (~35 min)
```
